---
title: "Project B (1)"
author: "Palle Villesen"
output: 
  html_document:
    code_folding: show
    theme: cosmo
    toc: yes
editor_options: 
  chunk_output_type: console
---

# Goals for this week 

* Basic questions about the datasets

* A simple plot refresher

* Use the shapiro.test() to test if gene expression is normally distributed. 

# First a bit about the Khan data

The data consists of a number of tissue samples corresponding to four distinct types of small round blue cell tumors.

For each tissue sample, 2308 gene expression measurements are available. 

This data were originally reported in: Khan J, Wei J, Ringner M, Saal L, Ladanyi M, Westermann F, Berthold F, Schwab M, Antonescu C, Peterson C, and Meltzer P. Classification and diagnostic prediction of cancers using gene expression profiling and artificial neural networks. Nature Medicine, v.7, pp.673-679, 2001. 

Here we will consider the subset of the `Khan` dataset where we focus on differences in gene expression between two cancer sub-types: 

+ Rhabdomyosarcoma aka group `2` 

+ Burkitt’s lymphoma aka group `4`

```{r}

library(tidyverse)
library(knitr) # required for nice table output
library(ggthemes) # required for nice colors

options(digits = 3) # number of digits printed by R default (vectors, data.frames, lists)
options(pillar.sigfig = 3) # number of digits printed by tibbles default.

set.seed(42) # to reproduce the example
theme_set(theme_minimal(base_size = 18)) # to get homogeneous figs

df_khan <- read_rds(r"(U17\khan_2vs4.tibble2.rds)")
```

# Warm up exercises

>Q1. How many samples do we have for each of the two cancer types?


```{r}
table(df_khan$cancer_group)

df_khan %>%
    group_by(cancer_group) %>%
    count()
```


2: 29
4: 25

---


>Q2. How many genes do we have in total?

```{r}
df_khan %>%
    select(-cancer_group) %>%
    ncol()
```

---

# One example to set the scene: The 11th gene

We focus on the numerical column `gene_11`

Let's practice again the tidyverse summaries by groups 

We first look at the sample size and the observed means

We use the infamous `kable()` to present results 

```{r}

df_khan %>%
    group_by(cancer_group) %>%
    summarize(
        n_obs = n(),
        mean_exp = mean(gene_11)
    ) %>%
    kable(col.names = c("Cancer group", "n", "Mean expression"))
```

>Q3. Use the sd() and sqrt() function to calculate the standard error of the mean for the two groups.


```{r}
df_khan %>%
    group_by(cancer_group) %>%
    summarize(
        n_obs = n(),
        SEMlow = mean(gene_11) - 2 * (sd(gene_11) / sqrt(n())),
        SEMhigh = mean(gene_11) + 2 * (sd(gene_11) / sqrt(n()))
    ) %>%
    kable(col.names = c("Cancer group", "n", "SEMlow", "SEMhigh"))
```

---

## Plotting a single gene

>Q4. Make a facetted histogram plot of gene 11.


```{r}
df_khan %>% ggplot(aes(x = gene_11)) +
    geom_histogram(bins = 10) +
    facet_wrap(~cancer_group)
```

---

Here we make a violin plot and add the jittered data.

```{r}

df_khan %>%
    ggplot(aes(x = cancer_group, y = gene_11)) +
    geom_violin() +
    geom_jitter(height = 0, width = 0.1, aes(color = cancer_group)) +
    ylab("Gene expression in gene 11 ") +
    xlab("Cancer type") +
    scale_color_colorblind() +
    theme(legend.position = "none") +
    NULL
```

## Checking assumptions

We want to check if our data are normally distributed. To do so we use the Shapiro-Wilk test (Chapter 13).

I will now show you a small example on how to use it. And then we want you to spend the rest of your time to get everything up and running.

```{r}

df_test <- tibble(
    x1 = c(
        4.33, 2.05, 1.93, 3.11, 3.35, 1.63, 2.61, 3.32, 2.15, 4.01,
        0.72, 2.18, 3.49, 3.93, 3.07, 2.44, 4.27, 1.37, 3.51, 2.31,
        3.31, 1.27, 4.64, 3.46, 1.31
    ),
    x2 = c(
        1.21, 2, 2.27, 0.38, 0.27, 2.37, 0.88, 0.72, 3.93, 0.23,
        1.66, 0.17, 0.1, 4.94, 0.5, 0.31, 0.64, 3.84, 2.01, 1.22,
        0.33, 8.12, 0.32, 2.92, 0.71
    )
)

shapiro.test(x = df_test$x1)

shapiro.test(x = df_test$x2)
```

>Q5. Which of the two variables x1 and x2 are NOT normally distributed?

x1 has a high probability to come from a normal distribution
x2 has a low probability of coming from a normal distribution (We can not reject our H0 hypothesis)

---

>Q6. What if you log transform the two variables?


```{r}
shapiro.test(x = log(df_test$x1, base = 4.20))
shapiro.test(x = log(df_test$x2, base = 4.20))
```

#x1
Shapiro-Wilk normality test

data:  log(df_test$x1, base = 4.2)
W = 0.9, p-value = 0.05

#x2
Shapiro-Wilk normality test

data:  log(df_test$x2, base = 4.2)
W = 1, p-value = 0.8

would u look at that :O


There are several different ways of log transforming in R. Your job is to try at least one of them and test the results of the shapiro.test() on x1 and x2 from before.

```{r}

pd <- tibble(
    x = seq(0, 10, length.out = 100),
    y = x,
    log_y = log(y),
    log2_y = log(y, base = 2), # or log2()
    log2_y2 = log2(y + 1),
    log10_y = log10(y)
) %>% pivot_longer(-x)

ggplot(pd, aes(x = x, y = value, color = name)) +
    geom_point()
```


---

# Checking assumptions of all the genes.

We now want you to test all genes and see if they look normally distributed.

We go step by step.

First we look at gene 11.

```{r}

ggplot(
    data = df_khan,
    mapping = aes(x = gene_11)
) +
    geom_histogram() +
    geom_vline(color = "#00ff37", xintercept = 1.00) +
    facet_wrap(~cancer_group, ncol = 1) +
    NULL
```

Then we extract group 2 and test for normality.

```{r}

df_khan_2 <- df_khan %>%
    filter(cancer_group == "2") %>%
    select(starts_with("gene"))


gene_values2 <- df_khan_2[, 11] %>% pull()
shapiro.test(gene_values2)
```
#Cancer group 2 - Gene 11
Shapiro-Wilk normality test

data:  gene_values
W = 1, p-value = 0.7



>Q7. Write the code to filter, extract and test group 4 - for gene 11.

```{r}

df_khan_4 <- df_khan %>%
    filter(cancer_group == "4") %>%
    select(starts_with("gene"))

gene_values4 <- df_khan_4[, 11] %>% pull()
shapiro.test(gene_values4)
```

shapiro.test(gene_values4)

#Cancer group 4 - Gene 11
Shapiro-Wilk normality test

data:  gene_values4
W = 1, p-value = 0.5

---

Now we use a `for()` loop and output the statistic we get every time

```{r}

set.seed(0)

results <- tibble(group2 = rep(NA, 8), group4 = NA)

df_khan_2 <- df_khan %>%
    filter(cancer_group == "2") %>%
    select(starts_with("gene"))



for (i in 1:8) {
    gene_values <- df_khan_2[, i] %>% pull()
    x <- shapiro.test(gene_values)
    results$group2[i] <- x$p.value
    cat("Finished gene ", i, "\n")
    flush.console()
}

results
```

>Q8. Rewrite the code so it tests group 2 and group 4 and do it for all genes.

```{r}

set.seed(0)

results <- tibble(group2 = rep(NA, 2308), group4 = rep(NA, 2308)) # 2308 is number of genes

df_khan_2 <- df_khan %>%
    filter(cancer_group == "2") %>%
    select(starts_with("gene")) # Makes the df with all genes of cancer type 2

df_khan_4 <- df_khan %>%
    filter(cancer_group == "4") %>%
    select(starts_with("gene")) # Makes the df with all genes of cancer type 4

for (i in 1:nrow(results)) {
    results$group2[i] <- shapiro.test(df_khan_2[, i] %>% pull())$p.value
    results$group4[i] <- shapiro.test(df_khan_4[, i] %>% pull())$p.value
    cat("Finished gene ", i, "\n")
    flush.console()
}

results
```

How many genes are NOT normally distributed in cancer  type 2?


```{r}

results %>%
    filter(group2 < 0.05) %>%
    select(group2) %>%
    count()
```

471 genes are not normally distributed in cancer type 2.

How many genes are NOT normally distributed in cancer  type 4?

```{r}

results %>%
    filter(group4 < 0.05) %>%
    select(group4) %>%
    count()
```


458 genes are not normally distributed in cancer type 4.

BONUS: How many did you expect? (remember that even Shapiro test can make false positives)



BONUS: are the p values correlated(?) (is it the same genes that are NOT normal in group 2 and group4)?

```{r}

results %>%
    filter(group4 < 0.05 & group2 < 0.05) %>%
    count()
```

```{r}
(137 * 2) / (471 + 458)
```

only 137 are both not normal. (about 30% of all not normals) seems weak.

---

