---
output:
  html_document:
        theme: readable
editor_options: 
  chunk_output_type: inline
---



```{r message=FALSE, warning=FALSE}

library(tidyverse)

options(digits = 7) # number of digits printed by R default (vectors, data.frames, lists)
options(pillar.sigfig = 7) # number of digits printed by tibbles default.
```

# World-wide life expectancy and income

## Background


The gapminder dataset provides information about income and life expectancy in all continents over the 60 years. 

Here the data was obtained from the gapminder R package and a subset of this data is made available as rds file. 

We used the following code to get the data and save it.



```{r}

library(gapminder)
data("gapminder")
summary(gapminder)

# Transforming data for the sake of the exam
years_to_keep <- c("1992", "2002")

gap10 <- gapminder %>%
    select(-c(pop, gdpPercap)) %>%
    filter(year %in% years_to_keep)

widegap10 <- gap10 %>%
    pivot_wider(names_from = year, values_from = lifeExp)

names(widegap10) <- c("country", "continent", "life_expec_1992", "life_expec_2002")

saveRDS(object = widegap10, file = "t1.gap_minder.rds")
```




The rds file `t1.gap_minder.rds` is made available below in R as the tibble `df`

```{r}

df <- readRDS("t1.gap_minder.rds")
head(df)

```

>Q1.1 How many countries have a reported life expectancy in 1992 ?

```{r}
length(df$life_expec_1992)
```


---

>Q1.2 How many Asian countries had a life expectancy of more than 60 years in 2002?

```{r}
df %>%
    filter(continent == "Asia" & life_expec_2002 > 60) %>%
    select(country) %>%
    dim()
```


---

>Q1.3 In how many countries did life expectancy increase from 1992 to 2002?

```{r}
df %>%
    filter(life_expec_1992 < life_expec_2002) %>%
    select(country) %>%
    dim()
```

---

