---
output:
  html_document:
        theme: readable
editor_options: 
  chunk_output_type: inline
---


## Background

```{r message=FALSE, warning=FALSE}

library(tidyverse)
options(digits = 7)        # number of digits printed by R default (vectors, data.frames, lists)
options(pillar.sigfig = 7) # number of digits printed by tibbles default.

```

```{r}

df <- read_rds("t3.fifa_results_data.rds")
head(df)
table(df$tournament)

```

>Q3.1 How many registered matches where Spain was playing at home do we have data for?

Hint: Note that in the dataset `df`, information on which team was considered the "home team" is recorded in the column `home_team` 
```{r}
sum(df$home_team=="Spain")

```

---

>Q3.2 What is the maximum number of goals scored by Spain when playing at home?

```{r}
df %>% filter(home_team=="Spain") %>% select(home_score)%>% max()

```




Hint1: again, by convention all matches where Spain is marked as `home_team`are considered to be played at home (although in some cups/ tournament the match was physically held somewhere outside Spain)

Hint2. The record is actually pretty high :-)


---

## A propobability model for the distribution of number of goals

Lets focus on matches where Spain played at home in the category `Friendly` matches 

(check that you have n = 187). 

>Q3.3  what is the mean of the number of goals scored by Spain in these 187 matches ?

```{r}
df %>% filter(home_team=="Spain" & tournament=="Friendly") %>% select(home_score)%>% colMeans()
```

---
