---
output:
  html_document:
        theme: readable
editor_options: 
  chunk_output_type: console
---

# Pain killer experiment

Imagine that we have tested the effect of weak painkillers in a randomized controlled trial.

The setup was simple:

We randomly divided people into two group: "Placebo" and "Drug".

First we gave them a small electric shock and they reported the pain level they experienced (0-100). 

Then they got a pill (placebo pill or active pill).

After waiting for 15 minutes, we gave them another small electric shock and they reported the pain level they experienced (0-100).

```{r message=FALSE, warning=FALSE}

library(tidyverse)

df <- read_tsv(file = r"(exam-prep\2020.ordinary_exam.june\t2.dataset.pain_experiment.tsv)")
```

>Q2.1 How many individuals were analyzed in this dataset?

>Q2.2 How many placebo pills did we give out?

>Q2.3 What was the mean change in pain level (change = After-Before) in the placebo group?

>Q2.4 What was the mean change in pain level (change = After-Before) in the drug group?

---------!begin suggested answer ------------------------

A few students calculate correctly but report a negative change as positive. (q2.3 and q2.4)

We have edited all your answers to positive values - and edited the answer key as well.


```{r}

df <- df %>% mutate(difference = After - Before)

answer <- tibble()
result <- nrow(df)


answer <- rbind(answer, tibble(
    question    = "1",
    true_answer = result,
    type        = "numeric",
    points      = 1,
    tolerance   = 0.1
))


result <- df %>%
    filter(group == "Placebo") %>%
    nrow()

answer <- rbind(answer, tibble(
    question    = "2",
    true_answer = result,
    type        = "numeric",
    points      = 1,
    tolerance   = 0.1
))


df2 <- df %>% filter(group == "Placebo")

answer <- rbind(answer, tibble(
    question    = "3",
    true_answer = abs(mean(df2$difference)),
    type        = "numeric",
    points      = 2,
    tolerance   = 0.01
))

df2 <- df %>% filter(group == "Drug")

answer <- rbind(answer, tibble(
    question    = "4",
    true_answer = abs(mean(df2$difference)),
    type        = "numeric",
    points      = 2,
    tolerance   = 0.01
))

answer
```
---------!end suggested answer ------------------------

>Q2.5 Test if the treatment reponse is different from 0 (no response) in the Placebo group, report the p value (two tailed test).

>Q2.6 Test if the treatment reponse in the Drug group is different than in the Placebo group. Report the p value (two tailed test).

---------!begin suggested answer ------------------------

```{r}

pd <- df %>% filter(group == "Placebo")

# wrong: data are paired - THIS IS NOT COEECT
(t.test(x = pd$After, y = pd$Before, paired = FALSE))


# Correct, data are paired
(t.test(x = pd$After, y = pd$Before, paired = TRUE))

# The paired test is identical to test if the difference is 0)
(x <- t.test(x = pd$difference))

answer <- rbind(answer, tibble(
    question    = "5",
    true_answer = x$p.value,
    type        = "numeric",
    points      = 3,
    tolerance   = 10 * x$p.value
))


# Test if the reponse in the Drug group is different than in the Placebo group. Report the p value (two tailed test).
pd1 <- df %>% filter(group == "Placebo")
pd2 <- df %>% filter(group == "Drug")

#
# Some students may use the placebo group mean response as null in a t.test.
# This is not correct - since it does NOT take sample size of the placebo group into account.
# Wrong
(x <- t.test(x = pd2$difference, mu = mean(pd1$difference)))

# Using formula
(x <- t.test(difference ~ group, data = df))

# Using two vectors
(x <- t.test(x = pd1$difference, y = pd2$difference))

answer <- rbind(answer, tibble(
    question    = "6",
    true_answer = x$p.value,
    type        = "numeric",
    points      = 3,
    tolerance   = 10 * x$p.value
))

answer
```

---------!end suggested answer ------------------------

>Q2.7 How large a proportion of all individuals had a pain level > 90 before treatment? 

Hint: Proportion is a number in the range [0;1]

Now we define a strong response as a decrease in pain level > 10.

>Q2.8 How large a proportion of individuals in the placebo group showed a strong reponse?

>Q2.9 How large a proportion of individuals in the drug group showed a strong reponse?

---------!begin suggested answer ------------------------

```{r}

# proportion

result <- sum(df$Before > 90) / nrow(df)

answer <- rbind(answer, tibble(
    question    = "7",
    true_answer = result,
    type        = "numeric",
    points      = 2,
    tolerance   = 0.01
))

pd1 <- df %>% filter(group == "Placebo")

result <- sum(pd1$difference < -10) / nrow(pd1)

answer <- rbind(answer, tibble(
    question    = "8",
    true_answer = result,
    type        = "numeric",
    points      = 2,
    tolerance   = 0.01
))

pd2 <- df %>% filter(group == "Drug")

result <- sum(pd2$difference < -10) / nrow(pd2)

answer <- rbind(answer, tibble(
    question    = "9",
    true_answer = result,
    type        = "numeric",
    points      = 2,
    tolerance   = 0.01
))

answer
```

## Hard coded information

```{r}

topic <- "t2" # If you change this to e.g. 2 - you should search & replace all >QX. with >QY.
topic_description <- "# Pain experiment"
file_description <- "pain_experiment"
```

## Saving exam files

```{r}

answer$question <- paste(topic, ".", answer$question, sep = "") # Reformat column

# Save answer sheet for students to fill out (including points)
x <- paste("answer$", answer$question, " = NA # (", answer$points, " points)", sep = "")

answer_filename <- paste("exam_answer_sheet.", topic, ".", file_description, ".txt", sep = "")
write_lines(topic_description, path = answer_filename)
write_lines(x, path = answer_filename, append = T)

# Save answer key for grading
key_filename <- paste("answer.key.", topic, ".", file_description, ".tsv", sep = "")
write_tsv(x = answer, path = key_filename)
```

---------!end suggested answer ------------------------

