---
output:
  html_document:
        theme: readable
editor_options: 
  chunk_output_type: console
---

# A really bad experiment

Imagine that I have a machine that records static noise from the atmosphere and convert it to two values (x and y).

On two different nights I recorded the following data (night A and B)

```{r message=FALSE, warning=FALSE}

library(tidyverse)

df <- read_tsv(file = r"(exam-prep\2020.ordinary_exam.june\t4.dataset.bad_data_permutation.tsv)")
```

>Q4.1 How many datapoints did I collect in total?

>Q4.2 How many datapoints did I collect in group A?

>Q4.3 What is the mean x value in group A?

>Q4.4 What is the absolute difference in means of x in the two groups?

Hint: abs(-3.5) = 3.5

---------!begin suggested answer ------------------------

Data points are confused. Some say x and y are two data points pr. observation. My bad.

We accept both answers (nrow() and nrow()*2) as correct in Q4.1 and Q4.2

Also, Q4.3 is reported with only 1 digit on screen (tidyverse problem/feature). So we accept this answer as well.

```{r}

answer <- tibble()
result <- nrow(df)

answer <- rbind(answer, tibble(
    question    = "1",
    true_answer = nrow(df),
    type        = "numeric",
    points      = 1,
    tolerance   = 0.1
))

answer <- rbind(answer, tibble(
    question    = "2",
    true_answer = sum(df$group == "A"),
    type        = "numeric",
    points      = 1,
    tolerance   = 0.1
))

df2 <- df %>% filter(group == "A")

answer <- rbind(answer, tibble(
    question    = "3",
    true_answer = mean(df2$x),
    type        = "numeric",
    points      = 2,
    tolerance   = 0.1
))

g1 <- df %>% filter(group == "A")
g2 <- df %>% filter(group == "B")
result <- abs(mean(g1$x) - mean(g2$x))

answer <- rbind(answer, tibble(
    question    = "4",
    true_answer = result,
    type        = "numeric",
    points      = 2,
    tolerance   = 0.01
))

answer
```
---------!end suggested answer ------------------------

Make a scatterplot of x and y for the two groups separatedly.

>Q4.5 What number is hidden in group A?

>Q4.6 What number is hidden in group B?

---------!begin suggested answer ------------------------

```{r}

ggplot(df, aes(x = x, y = y)) +
    geom_point() +
    facet_wrap(~group, ncol = 1) +
    NULL

answer <- rbind(answer, tibble(
    question    = "5",
    true_answer = 42,
    type        = "numeric",
    points      = 2,
    tolerance   = 0.01
))

answer <- rbind(answer, tibble(
    question    = "6",
    true_answer = 33,
    type        = "numeric",
    points      = 2,
    tolerance   = 0.01
))
```

---------!end suggested answer ------------------------

After looking at the data we can see that the x values are NOT normally distrited.

Despite the peculiar results above we try and look for other patterns in the data.

We specifically try to test if mean(x) is different in night A versus night B.

If we try and do a simple t.test (assuming equal variance in the two groups) we get:

```{r}

groupA <- df %>% filter(group == "A")
groupB <- df %>% filter(group == "B")

t.test(x = groupA$x, y = groupB$x, var.equal = T)
```

>Q4.7 What is the p value we get from this bad test?

---------!begin suggested answer ------------------------

This was a simple question. Just to see if you read the output correctly.

```{r}

groupA <- df %>% filter(group == "A")
groupB <- df %>% filter(group == "B")
x <- t.test(x = groupA$x, y = groupB$x, var.equal = T)

answer <- rbind(answer, tibble(
    question    = "7",
    true_answer = x$p.value,
    type        = "numeric",
    points      = 3,
    tolerance   = 0.01
))
```

---------!end suggested answer ------------------------

I really do not trust my test.

So I go ahead and make a permutation test using the following simple code (only 100 permutations)

```{r}

set.seed(0)
permuted <- rep(NA, 100)
for (i in 1:length(permuted)) {
    xvalues <- sample(df$x)
    g1 <- mean(xvalues[df$group == "A"])
    g2 <- mean(xvalues[df$group == "B"])
    permuted[i] <- abs(g1 - g2)
}

permuted
```


>Q4.8 How many of the 100 permutations were equal to or more extreme than the observed difference?

---------!begin suggested answer ------------------------

This is in order to test if you could correctly sum up and compare the numbers

```{r}

xvalues <- df$x
g1 <- mean(xvalues[df$group == "A"])
g2 <- mean(xvalues[df$group == "B"])
observed <- abs(g1 - g2)

set.seed(0)
permuted <- rep(NA, 100)
for (i in 1:length(permuted)) {
    xvalues <- sample(df$x)
    g1 <- mean(xvalues[df$group == "A"])
    g2 <- mean(xvalues[df$group == "B"])
    permuted[i] <- abs(g1 - g2)
}

result <- sum(permuted >= observed)

answer <- rbind(answer, tibble(
    question    = "8",
    true_answer = result,
    type        = "numeric",
    points      = 3,
    tolerance   = 0.1
))
```

---------!end suggested answer ------------------------

>Q4.9 What is the estimated p value based on 5000 permutations?

---------!begin suggested answer ------------------------

And here we tested if you remeber the "+1" and were able to change my code a little bit.

```{r}

xvalues <- df$x
g1 <- mean(xvalues[df$group == "A"])
g2 <- mean(xvalues[df$group == "B"])
observed <- abs(g1 - g2)

set.seed(0)
permuted <- rep(NA, 5000)
for (i in 1:length(permuted)) {
    xvalues <- sample(df$x)
    g1 <- mean(xvalues[df$group == "A"])
    g2 <- mean(xvalues[df$group == "B"])
    permuted[i] <- abs(g1 - g2)
}

pvalue <- (sum(permuted >= observed) + 1) / (length(permuted) + 1)

answer <- rbind(answer, tibble(
    question    = "9",
    true_answer = pvalue,
    type        = "numeric",
    points      = 3,
    tolerance   = 0.01
))
```

---------!end suggested answer ------------------------

---------!begin suggested answer ------------------------

## Hard coded information

```{r}

topic <- "t4" # If you change this to e.g. 2 - you should search & replace all >Q4. with >Q2.
topic_description <- "# Bad data and permutations"
file_description <- "bad_data_permutation"
```

## Answers table

```{r}

answer

```


## Saving exam files

```{r}

answer$question <- paste(topic, ".", answer$question, sep = "") # Reformat column

# Save answer sheet for students to fill out (including points)
x <- paste("answer$", answer$question, " = NA # (", answer$points, " points)", sep = "")

answer_filename <- paste("exam_answer_sheet.", topic, ".", file_description, ".txt", sep = "")
write_lines(topic_description, path = answer_filename)
write_lines(x, path = answer_filename, append = T)

# Save answer key for grading
key_filename <- paste("answer.key.", topic, ".", file_description, ".tsv", sep = "")
write_tsv(x = answer, path = key_filename)
```

---------!end suggested answer ------------------------

