---
title: "Project B (3)"
author: "Palle Villesen"
output: 
  html_document:
    code_folding: show
    theme: cosmo
    toc: yes
editor_options: 
  chunk_output_type: console
---

# Special code

The goal of this code is to run all tests needed for week 3.

We do the following: 

 * shapiro test all genes
 * t test all genes
 * wilcox test all genes
 * permutation test all genes

# Init

```{r}

library(tidyverse)
library(knitr) # required for nice table output
library(ggthemes) # required for nice colors
library(dqrng)

options(digits = 3) # number of digits printed by R default (vectors, data.frames, lists)
options(pillar.sigfig = 3) # number of digits printed by tibbles default.

set.seed(42) # to reproduce the example
theme_set(theme_minimal(base_size = 18)) # to get homogeneous figs
FILEPATHRES = r"(U19\all_test_results.rds)"
FILEPATHLOG = r"(U19\log.txt)"

df_khan <- read_rds(r"(U19\khan_2vs4.tibble2.rds)")
```

# Creating our output tibble

```{r}

results <- tibble(
    column = 2:ncol(df_khan),
    gene_name = names(df_khan)[column],
    shapiro2 = NA,
    shapiro4 = NA,
    log_fc = NA, # log2(fold_change)
    pval_ttest = NA, # for the p value from the t.test
    pval_wilcox = NA, # for the p value from the wilcox.test
    pval_perm = NA # for the p value from the permutation test
)

write_rds(x = results, file = FILEPATHRES)
```

# Shapiro test

```{r}

g2 <- df_khan %>%
    filter(cancer_group == "2") %>%
    select(starts_with("gene"))

starttime <- Sys.time()

for (i in 1:nrow(results)) {
    gene_name <- results$gene_name[i]
    gene_values <- g2 %>% pull(gene_name)
    x <- shapiro.test(gene_values)
    results$shapiro2[i] <- x$p.value
}

endtime <- Sys.time()

endtime - starttime


results

g4 <- df_khan %>%
    filter(cancer_group == "4") %>%
    select(starts_with("gene"))

starttime <- Sys.time()

for (i in 1:nrow(results)) {
    gene_name <- results$gene_name[i]
    gene_values <- g4 %>% pull(gene_name)
    x <- shapiro.test(gene_values)
    results$shapiro4[i] <- x$p.value
}

endtime <- Sys.time()

endtime - starttime

results

write_rds(x = results, file = FILEPATHRES)
```

# Fold change

The data are already log2() transformed so log2(g2/g4) is the same as:

log2(g2) - log2(g4).

```{r}

g2 <- df_khan %>%
    filter(cancer_group == "2") %>%
    select(starts_with("gene"))
g4 <- df_khan %>%
    filter(cancer_group == "4") %>%
    select(starts_with("gene"))

for (i in 1:nrow(results)) {
    gene_name <- results$gene_name[i]
    group2_values <- g2 %>% pull(gene_name)
    group4_values <- g4 %>% pull(gene_name)
    results$log_fc[i] <- mean(group2_values) - mean(group4_values)
}

results
```

# T test

```{r}

g2 <- df_khan %>%
    filter(cancer_group == "2") %>%
    select(starts_with("gene"))
g4 <- df_khan %>%
    filter(cancer_group == "4") %>%
    select(starts_with("gene"))

starttime <- Sys.time()

for (i in 1:nrow(results)) {
    gene_name <- results$gene_name[i]
    group2_values <- g2 %>% pull(gene_name)
    group4_values <- g4 %>% pull(gene_name)
    x <- t.test(x = group2_values, y = group4_values)
    results$pval_ttest[i] <- x$p.value
}

endtime <- Sys.time()

endtime - starttime

results
```

# Wilcox test

```{r}

g2 <- df_khan %>%
    filter(cancer_group == "2") %>%
    select(starts_with("gene"))
g4 <- df_khan %>%
    filter(cancer_group == "4") %>%
    select(starts_with("gene"))

starttime <- Sys.time()

for (i in 1:nrow(results)) {
    gene_name <- results$gene_name[i]
    group2_values <- g2 %>% pull(gene_name)
    group4_values <- g4 %>% pull(gene_name)
    x <- wilcox.test(x = group2_values, y = group4_values)
    results$pval_wilcox[i] <- x$p.value
}

endtime <- Sys.time()

endtime - starttime

results

write_rds(x = results, file = FILEPATHRES)
```

# Permutation test

## About the permutation test

The permutation test needs up to 1.000.000 permutations for each gene in order to get significant results after Bonferroni correction.

The code we used in week 2 could do 1 million permutations in ~10 seconds but it would take ~8 hours to finish. Your computer may be faster or slower - so your estimate may be different.

BUT both my and your computer is too slow for 2300*1.000.000 = 2.3 billion permutations.

So I have optimized it further.

## High speed adaptive permutation test of all genes

### Trick 1 - faster sample() function

Somebody made a sample() function that is faster.

I yes - I found it in stack overflow.

So we use that. It get's us down to ~8 seconds for 1 million permutations.

Still not fast enough.

### Trick 2 - stop permuting if the pvalue is not even close to significant. 

We keep an eye of extreme permutations and stop when we have "enough" extreme permutations to be quite certain of our p value estimate.

We don't care if we have 201/1000 (p ~ 0.201) or 201.345/1.000.000 (p ~0.201345) - we just need the 0.2 and could stop after 1000 permutations.

We DO care if we have 0/1000 after 1000. And only 1/100.000 - then we need more permutations.

So my little trick is to 

1. run 500 permutations
2. check if we have >20 extreme permutations
3. If not - continue.
4. If >20 stop and continue with next gene.

It saves a LOT of time.

The code below will permute up to 1.000.000 times pr gene (at least 500).

# Palles super optimized high speed permutation test.

```{r}

set.seed(0)

g2 <- df_khan %>%
    filter(cancer_group == "2") %>%
    select(starts_with("gene"))
g4 <- df_khan %>%
    filter(cancer_group == "4") %>%
    select(starts_with("gene"))

n1 <- nrow(g2)
n2 <- nrow(g4)

starttime <- Sys.time()

permute_ts <- rep(NA, 500)

for (j in 1:nrow(results)) { # for each gene
    gene_name <- results$gene_name[j]
    group2_values <- g2 %>% pull(gene_name)
    group4_values <- g4 %>% pull(gene_name)
    observed <- abs(mean(group2_values) - mean(group4_values)) # Observed difference
    n_extreme <- 0
    values <- c(group2_values, group4_values)
    nvalues <- length(values)
    for (i1 in 1:2000) { # We do at most 5000 chunks
        for (i2 in 1:500) { # each chunk is 200 permutations
            # permuted_values <- sample(values)                          # Default sample() (12 sec)
            permuted_values <- values[dqrng::dqsample.int(n = nvalues)] # high speed sample (8 sec)
            permute_ts[i2] <- abs(mean(permuted_values[1:n1]) - mean(permuted_values[(1 + n1):(n1 + n2)]))
        }
        n_extreme <- n_extreme + sum(permute_ts >= observed)
        # Check if we have >20 extreme permutations. If so break out of the for loop.
        if (n_extreme > 20) {
            break()
        }
    }
    n_permutations <- i1 * 500
    p_value <- (n_extreme + 1) / (n_permutations + 1)
    results$pval_perm[j] <- p_value
    cat(as.character(Sys.time()),
        gene_name, "done.",
        n_extreme, "extreme permutations found in", n_permutations, "permutations.",
        "P value ~", p_value, "\n",
        file = FILEPATHLOG, append = T
    )
}

endtime <- Sys.time()

endtime - starttime

results
```

# Quick plot of distributions

```{r}

ggplot(results, aes(x = log10(shapiro2))) +
    geom_histogram()

ggplot(results, aes(x = log10(shapiro4))) +
    geom_histogram()

ggplot(results, aes(x = log10(pval_ttest))) +
    geom_histogram()

ggplot(results, aes(x = log10(pval_wilcox))) +
    geom_histogram()

ggplot(results, aes(x = log10(pval_perm))) +
    geom_histogram()
```

# Save the results for the exercises.

```{r}

write_rds(x = results, file = FILEPATHRES)
```

