---
output:
  html_document:
        theme: readable
editor_options: 
  chunk_output_type: console
---


```{r message=FALSE, warning=FALSE}

library(tidyverse)

options(digits = 7)        # number of digits printed by R default (vectors, data.frames, lists)
options(pillar.sigfig = 7) # number of digits printed by tibbles default.

```

# World-wide life expectancy and income

## Background

The gapminder dataset provides information about income and life expectancy in all continents over the 60 years. 

Here the data was obtained from the gapminder R package and a subset of this data is made available as rds file. 

We used the following code to get the data and save it.

```r

library(gapminder)
data("gapminder")
summary(gapminder)

#Transforming data for the sake of the exam 
years_to_keep <-c("1992", "2002")

gap10 <- gapminder %>%
  select(-c(pop, gdpPercap))  %>%
  filter( year %in%  years_to_keep)
 
widegap10 <- gap10 %>%
  pivot_wider(names_from = year, values_from = lifeExp)
 
names(widegap10) <- c("country", "continent", "life_expec_1992", "life_expec_2002")

saveRDS(object = widegap10, file = "t1.gap_minder.rds")

```
The rds file `t1.gap_minder.rds` is made available below in R as the tibble `df`

```{r}

df <- readRDS("t1.gap_minder.rds")
head(df) 

```

>Q1.1 How many countries have a reported life expectancy in 1992 ?


---

>Q1.2 How many Asian countries had a life expectancy of more than 60 years in 2002?


---

>Q1.3 In how many countries did life expectancy increase from 1992 to 2002?


---

>Q1.4 What proportion of countries located in Asia had a life expectancy exceeding 60 in 2002

NOTE: A proportion should be reported as a number between 0 and 1 [0, 1]. So 42% should be written as 0.42 or 42/100.


---

## Did life expectancy go up in Asia from 1992 to 2002?

You will now test if life expectancy went up in Asia from 1992 to 2020.

Under the null hypothesis ($H_0$) of no change in life expectancy, we expect that the values of 2002 have equal probability to go up or down relative to 2019.

We have the alternative hypothesis that life expectancy is going up (i.e. it is a one tailed test).

Count the number of countries in Asia where the life expectancy is higher in 2002 compared to 1992. 

Calculate the probability under $H_0$ that as many countries as observed or more have a greater life expectancy in 2002 relative to 1992. (this is a one tailed probability).

>Q1.5 What is the calculated probability? (this also the one tailed p value)


---
