---
output:
  html_document:
        theme: readable
editor_options: 
  chunk_output_type: console
---

# World-wide happinees

## Background

The World Happiness Report is an annual publication of the United Nations Sustainable Development Solutions Network. Here the data was obtained from Kaggle: 
<https://www.kaggle.com/mathurinache/world-happiness-report>.
A subset of the data (`happy_integrated_2019_2020.rds`) is made available below as the tibble `df`

```{r message=FALSE, warning=FALSE}

library(tidyverse)

options(digits = 7) # number of digits printed by R default (vectors, data.frames, lists)
options(pillar.sigfig = 7) # number of digits printed by tibbles default.

df <- readRDS(r"(exam-prep\2021.ordinary_exam.june\t1.happy_integrated_2019_2020.rds)")
head(df)
```

>Q1.1 How many countries have a reported happiness score in both 2019 and 2020 ?

```{r}
nrow(df)
```
147
---

>Q1.2 How many Western European countries have a combined happiness score exceeding 7 in 2019/2020 ?

Hint: Make a new combined score variable `df$combined` that is defined as the mean happiness score in 2019 and 2020.

```{r}
df <- df %>% mutate(combined = (score_2020 + score_2019) / 2)

df %>%
    filter(region == "Western Europe") %>%
    count(combined > 7)
```
12
---

>Q1.3 In how many countries did happiness increase from 2019 to 2020?

```{r}
df %>% count(score_2019 < score_2020)
```
97
---

>Q1.4 What proportion of countries located in Sub-Saharan Africa had a happiness score exceeding 6 in 2020

NOTE: A proportion should be reported as a number between 0 and 1 [0, 1]. So 42% should be written as 0.42 or 42/100.

```{r}
df %>%
    filter(region == "Sub-Saharan Africa") %>%
    select(score_2020) %>%
    count(.$score_2020 > 6)
```
1/39
---

## Did happiness go up in Sub-Saharan Africa from 2019 to 2020?

You will now test if Happiness went up in Sub-Saharan Africa from 2019 to 2020.

Under the null hypothesis ($H_0$) of no change in happiness, we expect that the scores of 2020 have equal probability to go up or down relative to 2019.

We have the alternative hypothesis that happiness is going up (i.e. a one tailed test).

Count the number of countries in Sub-Saharan Africa where the happiness score is higher in 2020 compared to 2019. 

Calculate the probability under $H_0$ that as many countries as observed or more have a greater happiness in 2020 relative to 2019. (this is a one tailed probability).

>Q1.5 What is the calculated probability? (this also the one tailed p value)

```{r}

df %>%
    filter(region == "Sub-Saharan Africa") %>%
    count(score_2020 > score_2019)
true <- 27
false <- 12
binom.test(x = 27, n = 39, p = 0.5, alternative = "greater")
```
---
p-value = 0.01185
>Q1.6 Use the appropriate type of t-test to test whether happiness has gone up or down in 2020 relative to 2019 in Sub-Saharan Africa.

We assume that all the assumptions of the t-test are met.

```{r}

df_ssa <- df %>%
    filter(region == "Sub-Saharan Africa")
t.test(x = df_ssa$score_2020, y = df_ssa$score_2019, paired = T)
```
p-value = 0.023
---

>Q1.7: What is your conclusion when comparing both tests?

1. We can't reject H0 in Q.1.6, so there is no significant trend in happiness from 2019 to 2020.
2. The binomial test is better than a t-test so the p value should be lower in Q.1.5 than in Q 1.6
3. It's significant at the 0.05 level using both tests, so we conclude that happiness is going up. 
4. The t test in Q1.6 is not significant at the 0.05 level so we should not trust the test based on counts in Q.1.5 
5. Both test suggest happiness goes up in this region, but it is unlikely that there is a trend because world wide happiness seems to clearly go down in more than 70% of the countries.

The answer should be given as a single number: 1, 2, 3..

3

---

## Plotting from year to year

Now we want you to use your plotting skills to answer the following:

>Q1.8: If you plot 2019 vs. 2020, how would you describe Western Europe when compared to other regions:


```{r}
df %>% ggplot(aes(x = score_2019, y = score_2020)) +
    geom_line() +
    facet_wrap(~region)
```

1. The data are very spread out around a line with slope 1
2. The data are very spread out around a line that is nearly flat
3. The data are very close to a line with slope 1
4. The data are very close to a line that is nearly flat
5. The data are not linear but more exponential looking
6. The data show a negative trend (i.e. a clear negative slope)

The answer should be given as a single number: 1, 2, 3..
3

---

## Looking at changes

>Q1.9: Use your plotting skills to answer the following.

Hint: one of the years on X, the change on Y.

```{r}
df %>% ggplot(aes(x = score_2019, y = score_2020 - score_2019, color = region)) +
    geom_point()
```

1. Generally, happiness changed most for the countries with higher happiness scores in 2019.
2. Generally, happiness changed most for the countries with lower happiness scores in 2019.
3. There is no trend between happiness scores in 2019 and the change in happiness.
4. The country which improved the most is from Latin America and Caribbean.
5. The country which worsened the most is from Sub-Saharan Africa
6. Generally, the world became unhappier in 2020 than 2019.

The answer should be given as a single number: 1, 2, 3..

2


---

>Q1.10: Which region improved most? 

We define "improved most" as the region where the highest proportion of countries increased in happiness from 2019 to 2020. You can answer this in several ways (plotting or calculating it)

1. South Asia
2. Central and Eastern Europe
3. Middle East and North Africa
4. Latin America and Caribbean
5. Commonwealth of Independent States
6. North America and ANZ
7. Western Europe
8. Sub-Saharan Africa
9. Southeast Asia
10. East Asia

The answer should be given as a single number: 1, 2, 3..

```{r}
df %>%
    group_by(region) %>%
    summarise(
        up = sum(score_2020 > score_2019),
        n = n(),
        p = up / n
    ) %>%
    arrange(desc(p))
```

---


