---
output:
  html_document:
        theme: readable
editor_options: 
  chunk_output_type: console
---

# Births in Denmark

## Background

The use of IVF (in vitro fertilization) has changed dramatically in Denmark. 

Over the years, the implantation success has increased. At some point this meant that implanting three eggs nearly always resulted in either twins or triplets.

You will work on a dataset with the number of births of different types for different years.

The type can be "Case of single" (1 child), "Case of twins" (2 children) and so on. 

The data were downloaded directly from https://www.statistikbanken.dk/fod8 using R and are made available as `t3.birth_data.tsv`.

```{r message=FALSE, warning=FALSE}

library(tidyverse)
options(digits = 7)        # number of digits printed by R default (vectors, data.frames, lists)
options(pillar.sigfig = 7) # number of digits printed by tibbles default.

df <- read_tsv("t3.birth_data.tsv")

```

>Q3.1 How many different years have we collected data for?


---

>Q3.2 What is the highest number of babies ever born in Denmark at the same time?

Hint: If the most children ever born in a single event were twins (i.e. 2 children born at the same time), then the answer would be 2.


---

>Q3.3 What year was the first quadruplets (firlinger) recorded ?


---

>Q3.4 How many children were born as twins in the period 2000-2018 (both years included)?


---

>Q3.5 How many children were born as triplets in the period 2000-2018 (both years included)?


---

>Q3.6 How many children were born as quadruplets (across all years)?


---

## Testing different years

IVF treatment (with the horrible Danish name "reagensglasbørn") was invented in the UK in 1978, and the first baby ever recorded where the fertilization happened outside the body was Louise Brown. 

The first IVF baby was born in Denmark in 1983. So we know that all babies born before 1982 were the result of internal fertilization.

We know that IVF treatment often leads to multiple babies (because more than one egg is implanted). Now we want to compare three time periods: 1978-1982 (5 years before IVF), 1983-1987 (early IVF) and 2014-2018 (last 5 years).

Please note, that IVF treatment was probably rare in the 1980s and now it's really common. In 2019 we performed ~12.000 IVF treatments and ~2500 IVF children were born. (Source: https://sundhedsdatastyrelsen.dk/da/tal-og-analyser/analyser-og-rapporter/andre-analyser-og-rapporter/assisteret-reproduktion)

First, we reformat the data so we pool so a birth is either a single birth (1 child), twins (2 children) or more than two children (3+ or "flerfoldsfødsler" in Danish). 

This is done here:

```{r}

(df2 <- df %>% 
  mutate(type2 = ifelse(type=="Cases of single", "1", "3+")) %>%
  mutate(type2 = ifelse(type=="Cases of twins", "2", type2)) %>%
  group_by(year,type2) %>%
  summarise(n=sum(n)) %>%
  ungroup()
)

```

Now the rest is up to you. You must calculate the numbers for each period and do the correct testing.

You can NOT use fisher.test() for this, since the numbers are too high.

So you should use a chi square test to test if there is an association between period and birth type.

Hint: How to create a table from numbers - or from a tibble with counts.

```{r}

# Build a table from numbers
x <- matrix(c(20,69,16,157),ncol=2,byrow=FALSE)
colnames(x) <- c("Infected","Uninfected")
rownames(x) <- c("Multiple", "Single")
x

# From table to tibble and back
(x2 <- as_tibble(x) %>% mutate(parasites = rownames(x)))
(x3 <- x2 %>% pivot_longer(-parasites)) # counts
(x4 <- x3 %>% uncount(value)) # one row pr. observation
x4 %>% group_by(parasites, name) %>% count() # Check numbers are still correct

# From tibble to table
table(x4$parasites, x4$name)

```

>Q3.7: What is the p value when you compare 1978-1982 with 1983-1987?

H0: There is no association between the period and the type of births  
Ha: There is an association between the period and the type of births  


---

>Q3.8: What is your conclusion if the p value is 0.06 for this test?

1. We can't reject H0, so there is no association at all.
2. It's not significant, but it's pretty close, so there could be an association.
3. It's significant, so we are sure there is an association
4. It's around the significance threshold, so we can't say anything.
5. It's significant, so it's unlikely that there is no association.

The answer should be given as a single number: 1, 2, 3..


---

>Q3.9: What is the p value when you compare 1978-1982 with 2014-2018?

H0: There is no association between the period and the type of births  
Ha: THere is an association between the period and the type of births  


---

>Q3.10: What is your overall conclusion from the data and test results?

I define "multi births" as twins, triplets, quadruplets etc.

1. IVF treatment gave rise to more multi-births.
2. IVF treatment didn't influence the number of multi-births.
3. IVF treatment only influenced multi births in the 1980s.
4. IVF treatment resulted in fewer multibirths in the 2010-s.
5. Our two tests gave opposite results, I can't conclude anything

The answer should be given as a single number: 1, 2, 3..


---


