---
output:
  html_document:
        theme: readable
editor_options: 
  chunk_output_type: console
---


```{r message=FALSE, warning=FALSE}

library(tidyverse)

options(digits = 7)        # number of digits printed by R default (vectors, data.frames, lists)
options(pillar.sigfig = 7) # number of digits printed by tibbles default.

```

# World-wide life expectancy and income

## Background

The gapminder dataset provides information about income and life expectancy in all continents over the 60 years. 

Here the data was obtained from the gapminder R package and a subset of this data is made available as rds file. 

We used the following code to get the data and save it.

```r

library(gapminder)
data("gapminder")
summary(gapminder)

#Transforming data for the sake of the exam 
years_to_keep <-c("1992", "2002")

gap10 <- gapminder %>%
  select(-c(pop, gdpPercap))  %>%
  filter( year %in%  years_to_keep)
 
widegap10 <- gap10 %>%
  pivot_wider(names_from = year, values_from = lifeExp)
 
names(widegap10) <- c("country", "continent", "life_expec_1992", "life_expec_2002")

saveRDS(object = widegap10, file = "t1.gap_minder.rds")

```
The rds file `t1.gap_minder.rds` is made available below in R as the tibble `df`

```{r}

df <- readRDS("t1.gap_minder.rds")
head(df) 

```

>Q1.1 How many countries have a reported life expectancy in 1992 ?


---

>Q1.2 How many Asian countries had a life expectancy of more than 60 years in 2002?


---

>Q1.3 In how many countries did life expectancy increase from 1992 to 2002?


---

>Q1.4 What proportion of countries located in Asia had a life expectancy exceeding 60 in 2002

NOTE: A proportion should be reported as a number between 0 and 1 [0, 1]. So 42% should be written as 0.42 or 42/100.


---

## Did life expectancy go up in Asia from 1992 to 2002?

You will now test if life expectancy went up in Asia from 1992 to 2020.

Under the null hypothesis ($H_0$) of no change in life expectancy, we expect that the values of 2002 have equal probability to go up or down relative to 2019.

We have the alternative hypothesis that life expectancy is going up (i.e. it is a one tailed test).

Count the number of countries in Asia where the life expectancy is higher in 2002 compared to 1992. 

Calculate the probability under $H_0$ that as many countries as observed or more have a greater life expectancy in 2002 relative to 1992. (this is a one tailed probability).

>Q1.5 What is the calculated probability? (this also the one tailed p value)


---

>Q1.6 Use the appropriate type of t-test to test whether life expectancy has changed significantly from 1992 to 2002 in Asia. Report the p value as your answer.

We assume that all the assumptions of the t-test are met.


---

>Q1.7: What is your conclusion when comparing both tests?

1. We can't reject H0 in Q.1.6, so there is no significant trend in life expectancy from 1992 to 2002.
2. The binomial test is better than a t-test so the p value should be lower in Q.1.5 than in Q 1.6
3. It's significant at the 0.05 level using both tests, so we conclude that life expectancy is going up. 
4. The t test in Q1.6 is not significant at the 0.05 level so we should not trust the test based on counts in Q.1.5 
5. Both test suggest life expectancy does not go up in this region, but it is unlikely that there is a trend because world wide expectancy seems to clearly go down in more than 70% of the countries.

The answer should be given as a single number: 1, 2, 3..


---

## Plotting from year to year

Now we want you to use your plotting skills to answer the following:

>Q1.8: If you plot 2002 vs. 1992 how would you describe the Americas when compared to other regions:

1. The data are very spread out around a line with slope 1
2. The data are very spread out around a line that is nearly flat
3. The data are close to a line with slope 1
4. The data are close to a line that is nearly flat
5. The data are not linear but more exponential looking
6. The data show a negative trend (i.e. a clear negative slope)

The answer should be given as a single number: 1, 2, 3..


---

## Looking at changes

>Q1.9: Use your plotting skills and find the correct statement below.

Hint: one of the years on X, the change on Y.

1. Generally, life expectancy increased most for countries where life expectancy was lowest (below 55) scores in 1992.
2. Generally, life expectancy improved most for the countries with high life expectancy scores (eg above  65) in 1992.
3. There is no trend between life expectancy scores in 1992 and the change in life expectancy over 10 years
4. The country which improved the most is from Americas
5. The country which worsened the most is from Asia
6. Generally, life expectancy increased in Africa in 2002 relative to 1992.

The answer should be given as a single number: 1, 2, 3..


---

>Q1.10: Which continent improved most? 

We define "improved most" as the continent where the highest proportion of countries increased in life expectancy from 2019 to 2020. You can answer this in several ways (plotting or calculating it). We exclude Oceania from the calculation as only 2 countries are reported for this continent

1. Asia
2. Europe
3. Africa
4. Americas 
5. Antartica

The answer should be given as a single number: 1, 2, 3..


---


