---
output:
  html_document:
        theme: readable
editor_options: 
  chunk_output_type: console
---


## Background

```{r message=FALSE, warning=FALSE}

library(tidyverse)
options(digits = 7)        # number of digits printed by R default (vectors, data.frames, lists)
options(pillar.sigfig = 7) # number of digits printed by tibbles default.

```

```{r}

df <- read_csv("t3.fifa_results_data.csv")
head(df)
table(df$tournament)

```

>Q3.1 How many registered matches where Spain was playing at home do we have data for?

Hint: Note that in the dataset `df`, information on which team was considered the "home team" is recorded in the column `home_team` 

---------!begin suggested answer ------------------------

```{r}

answer_list <- list()

df_dk <- df %>%
  filter(home_team == "Spain")

(result1 <- nrow(df_dk))

answer_list[["1"]] <- tibble(
  true_answer = result1, 
  type        = "numeric",
  points      = 1,
  tolerance   = 0.1)

rm(df_dk, result1)

```

---------!end suggested answer ------------------------

---

>Q3.2 What is the maximum number of goals scored by Spain when playing at home?

Hint1: again, by convention all matches where Spain is marked as `home_team`are considered to be played at home (although in some cups/ tournament the match was physically held somewhere outside Spain)

Hint2. The record is actually pretty high :-)

---------!begin suggested answer ------------------------

```{r}

df_home <- df %>%
  filter(home_team == "Spain")  

summary(df_home)

(result2 <- max(df_home$home_score))

answer_list[["2"]] <- tibble(
  true_answer = result2, 
  type        = "numeric",
  points      = 1,
  tolerance   = 0.1)

qplot(df_home$home_score)

rm(df_home, result2)

```

---------!end suggested answer ------------------------

---

## A propobability model for the distribution of number of goals

Lets focus on matches where Spain played at home in the category `Friendly` matches 

(check that you have n = 187). 

>Q3.3  what is the mean of the number of goals scored by Spain in these 187 matches ?

---------!begin suggested answer ------------------------

```{r}

df2 <- df %>%
  filter(home_team == "Spain") %>%
  filter(tournament == "Friendly")
           
dim(df2)

(result <- mean(df2$home_score))

var(df2$home_score)

answer_list[["3"]] <- tibble(
  true_answer = result, 
  type        = "numeric",
  points      = 1,
  tolerance   = 0.01)

rm(df2)

```

---------!end suggested answer ------------------------

---

>Q3.4 What is the probability that Spain scores 6 goals or more in a friendly match at home ?

Hint: If we assume that goals have a constant probability of occurring per unit time during a football match, then we expect the number of goals in a match scored by a team to follow a Poisson probability distribution.

---------!begin suggested answer ------------------------

We assume that scores, $S$, follow a Poisson distribution with a mean X.

Then our best estimate of X is the observed mean in Spain's friendly matches at home.

Using that, we can then calculate  P[S >= 6]

```{r}

df2 <- df %>%
  filter(home_team == "Spain") %>%
  filter(tournament == "Friendly")
           
X <- mean(df2$home_score)

# Summing by hand, 6,7...up to 100 goals
probs <- dpois(x = 6:100, lambda = X) # the observed mean
sum(probs)

# We can also use ppois()
1-ppois(q = 5,lambda = X, lower.tail = T ) # 1 - lower tail
ppois(q = 5,lambda = X, lower.tail = F )   # upper tail

(result <- ppois(q = 5,lambda = X, lower.tail = F ))

answer_list[["4"]] <- tibble(
  true_answer = result, 
  type        = "numeric",
  points      = 2,
  tolerance   = 0.001)

rm(df2,X,probs,result)

```

---------!end suggested answer ------------------------

---

>Q3.5 What is the probability of Spain not scoring any goal during the first 45 mins of a friendly match at home ?

Hint: Here, we assume that goals have a constant probability of occurring per unit time. And just to be clear, we still assume that a football game last 2*45 = 90 mins. 
(we do not worry about the fact some matches might have involved overtime)

---------!begin suggested answer ------------------------

Under the Poisson assumption, we can "slice" the scores into smaller time intervals. So the mean of the first 45 minutes must be exactly 0.5 * mean(90 minutes).

So for half of a game we adjust the mean accordingly (to 0.5 og the value before).

```{r}

df2 <- df %>%
  filter(home_team == "Spain") %>%
  filter(tournament == "Friendly")
           
X <- 0.5*mean(df2$home_score)

(result <- dpois(x = 0 ,lambda = X))

answer_list[["5"]] <- tibble(
  true_answer = result, 
  type        = "numeric",
  points      = 3,
  tolerance   = 0.001)

```

---------!end suggested answer ------------------------

---

## Poisson distributed goals?

If the number of goals scored by a team is Poisson distributed (our null hypothesis), we expect the variance to mean of scores to be close to 1. 

Now we will focus on Spain playing in "FIFA World Cup qualification" at home.

>Q3.6 Calculate the variance to mean ratio of goals scored when Spain is playing in "FIFA World Cup qualification" at home.

Hint: just as help to check your consistency in data manipulation,  there where 57 matches where Spain was the home team for that category of `tournament`.

---------!begin suggested answer ------------------------

```{r}

table(df$tournament)

dfspain <- df %>%
  filter(home_team == "Spain") %>%
  filter(tournament == "FIFA World Cup qualification")

dim(dfspain)

(obsVtoM <- var(dfspain$home_score)/ mean(dfspain$home_score))

answer_list[["6"]] <- tibble(
  true_answer = obsVtoM, 
  type        = "numeric",
  points      = 3,
  tolerance   = 0.001)

```

---------!end suggested answer ------------------------

---

## Simulating datasets

Here is a small R code snippet to simulate distribution of that statistic using pseudo datasets of $n=50$ observations from a Poisson with mean  $\lambda = 1.2$ .

```{r}

var_to_mean <- rep(NA, 100) # A vector for storing the ratios simulated under H0

set.seed(0) # random seed init
for (i in 1:100) {
  pseudo_data <- rpois(n = 50, lambda = 1.2)
  var_to_mean[i] <- var(pseudo_data)/mean(pseudo_data)
}

summary(var_to_mean)

```

Adapt the code above to generate the distribution using 10.000 pseudo datasets of the same size as the observed Spanish data under the null hypothesis that the data is Poisson distributed. 

>Q3.7 Calculate and report the probability that the variance to mean ratio is equal or above the observed value in the actual data for Spain playing at home in the "FIFA World Cup qualification"?

---------!begin suggested answer ------------------------

```{r}

sample_size = nrow(dfspain)
X <- mean(dfspain$home_score)

var_to_mean <- rep(NA, 10000) # A vector storing the ratios simulated under H0
set.seed(42)
for (i in 1:10000) {
  pseudo_data <- rpois(n = sample_size, lambda = X)
  var_to_mean[i] <- var(pseudo_data)/mean(pseudo_data)
}

sum(var_to_mean >= obsVtoM)

# The +1 or not do not change the result. 
# We can debate whether you should have the +1 or not when we have 10000 simulations.
(result <- (1+sum(var_to_mean >= obsVtoM)) / (10000 + 1) )

answer_list[["7"]] <- tibble(
  true_answer = result, 
  type        = "numeric",
  points      = 3,
  tolerance   = 0.04) #that is 5 times bigger than the variation observed from simulation to simulation

```

---------!end suggested answer ------------------------

---

>Q3.8: What is your conclusion for the p value of this test (if you use 0.05 as significance threshold)?

1. We can't reject H0, so the Poisson model is probably correct.
2. The p value is just below 0.05 so the Poisson model is probably incorrect but we need more data.
3. No simulation produced a dataset with variance to mean ratio as extreme as the observed. So we can say that p < 0.001.
4. It's around the significance threshold, so we can't say anything.

The answer should be given as a single number: 1, 2, 3..

---------!begin suggested answer ------------------------

1. correct 
2. pvalue is much higher than 0.05
3. no
4  no

```{r}

(result <- 1)

answer_list[["8"]] <- tibble(
  true_answer = result, 
  type        = "numeric",
  points      = 1,
  tolerance   = 0.1)

```

---------!end suggested answer ------------------------

---

---------!begin suggested answer ------------------------

## Hard coded information

```{r}

topic             <- "t3"  # If you change this to e.g. 2 - you should search & replace all >Q3. with >Q2.
topic_description <- "# Footbal goals"
file_description  <- "football_goals"

```

## Answers table

```{r}

(
answer <- bind_rows(answer_list) %>%
  mutate(question = paste(topic, ".", names(answer_list), sep="")) %>%
  select(question, everything())
)

```

## Saving exam files

```{r}

# Save answer sheet for students to fill out (including points)
x <- paste("answer$", answer$question, " = NA # (",answer$points," points)",sep="")

answer_filename <- paste("exam_answer_sheet.",topic, ".", file_description, ".txt", sep="")
write_lines(topic_description, answer_filename)
write_lines(x, answer_filename, append = T)

# Save answer key for grading
key_filename <- paste("answer.key.",topic, ".", file_description, ".tsv", sep="")
write_tsv(answer, key_filename)

```

---------!end suggested answer ------------------------

