---
output:
  html_document:
        theme: readable
editor_options: 
  chunk_output_type: console
---

# Drugs in sewage water in European cities

The European Centre for Drugs and Drug Addiction is monitoring drug use in European cities.

Analyzing communal waste waters for drugs and their metabolic products in order to estimate their consumption in the community is a developing field, involving scientists working in different research areas, including analytical chemistry, physiology and biochemistry, sewage engineering, spatial epidemiology and statistics, and conventional drug epidemiology.

The collaboration have collected data from several drugs over the years from various cities in Europe. 

Data are originally found here: https://www.emcdda.europa.eu/topics/wastewater

The measured concentrations are given as X mg/1000 persons/day. So it is independent of city size. The data provided here is a summary of the average concentration over all days (Monday-Sunday) measured in a time window for a given year. 

If a city didn't collect any data a given year, the value is indicated as NA (= missing data).


```{r message=FALSE, warning=FALSE}

library(tidyverse)

options(digits = 7) # number of digits printed by R default (vectors, data.frames, lists)
options(pillar.sigfig = 7) # number of digits printed by tibbles default.

df <- read_rds(r"(Examen\t1.drugs_in_europe_data.rds)")

head(df)
```

## >Q1.1 How many different drugs do we have data for?

```{r}
n_distinct(df$metabolite)
```


## >Q1.2 How many different cities have a concentration > 0 in 2021 for amphetamine? 

```{r}
df %>%
    filter(metabolite == "amphetamine" & year_2021 > 0) %>%
    nrow()
```

## >Q1.3 What is the observed mean consumption of cocaine in Austria (AT) in 2021?

You may need to remove missing values when summarising. 

Many R functions have the `na.rm=T/F` parameter which can be used.

Example of using `na.rm=T/F` parameter in the `sum()` function.

```{r}

tibble(x = c(1, 2, 3, NA)) %>% summarise(median = sum(x)) # default is na.rm=F

tibble(x = c(1, 2, 3, NA)) %>% summarise(median = sum(x, na.rm = T))
```


```{r}
df %>%
    filter(country == "AT" & metabolite == "cocaine") %>%
    pull(year_2021) %>%
    mean(na.rm = T)
```


## >Q1.4 In how many cities were the measured cocaine consumption > 100 mg pr. 1000 people pr. day in 2021? 

```{r}
df %>%
    filter(metabolite == "cocaine" & year_2021 > 100) %>%
    nrow()
```
## >Q1.5 In how many cities did the measured cocaine consumption increase in 2021 relative to 2019? 

```{r}
df %>%
    filter(metabolite == "cocaine" & year_2021 > year_2019) %>%
    nrow()
```
# A theoretical MDMA example

NOTE: This is a purely theoretical question. You should NOT use the dataset to answer this.

## >Q1.6 If we assume that every city have equal probability (50:50) of going up/down in MDMA consumption from 2019 to 2021, what is the probability of observing "just by chance" precisely 40 out of 65 cities going up in consumption.

```{r}


dbinom(x = 40, size = 65, prob = 0.5)
```

Now let's assume that we actually did observe that 40 cities out of 65 went up in MDMA consumption.

Can you test the hypothesis:

H0: The probability of increased consumption is 0.5
Ha: The probability of increased consumption is NOT 0.5

## >Q1.7 Report the p value for the adequate test.

```{r}
binom.test(x = 40, n = 65, p = 0.5)
```

## >Q1.8 What do you conclude from your test results?

Answer should be a number indicating one of the following statements:

1. In more than half of the cities surveyed, MDMA consumption went up in 2021, relative to 2019. The trend is not significant (at the alpha=0.05 level).

2. In more than half of the cities surveyed, MDMA consumption went up in 2021, relative to 2019. The trend is strongly significant (at the alpha=0.05 level).

3. Mean MDMA consumption in Europe was significantly higher in 2021 relative to 2019. 

4. Mean MDMA consumption in Europe was significantly lower in 2021 relative to 2019. 

5. None of the above.

###
dataen er fra et hypotetisk H0 og er ikke surveyed, såehhhh. Dog er der en ikke significant trend op.
1 hvis "surveyed" er det i rent faktisk mener, ellers 5


# Back to the dataset

Using the city as an observation unit, compare the mean consumption of amphetamine in 2021 and 2019 using the appropriate type of test for comparing means.

You should start by checking if assumptions of normal data are met.

Then choose the correct test and

## >Q1.9  Report the p value of the test. 

```{r}
df2 <- df %>%
    filter(metabolite == "amphetamine") %>%
    select(city, year_2021, year_2019)

shapiro.test(df2$year_2021)
shapiro.test(df2$year_2019)

# dataen er altså ikke normalfordelt

wilcox.test(df2$year_2021, df2$year_2019)


# sanitytest for tolerance
t.test(df2$year_2021, df2$year_2019)
# seems about right. #paired seems wonky with na data.
```

Our null hypothesis: the mean consumption were the same in the two years. 

Our alternative hypothesis: the mean consumption were different in the two years.


## >Q1.10 What do you conclude from your test results?

Answer should be a number indicating one of the following statements:

1. I could not reject that data were normal, so I used t.test and found the result to be NOT significant.
2. I could not reject that data were normal, so I used t.test and found the result to be significant (p < 0.05)
3. I could not reject that data were normal, so I used t.test and found the result to be significant (p < 0.01)
4. I rejected that data were normal, so I used wilcox.test and found the result to be NOT significant.
5. I rejected that data were normal, so I used wilcox.test and found the result to be significant (p < 0.05)
6. I rejected that data were normal, so I used wilcox.test and found the result to be significant (p < 0.01)
7. None of the above.





