---
title: "Project B (2)"
author: "Palle Villesen"
output: 
  html_document:
    code_folding: show
    theme: cosmo
    toc: yes
editor_options: 
  chunk_output_type: console
---

# Goals for this week 

* Run t test, non-parametric test on a single gene.

* Run permutation test on a single gene.

* Optimize the running speed of the permutation test (code provided)

* Run t test & non-parametric test on all genes (and get the p values)

* Summarise your findings

# A refresher about the Khan data

The data consists of a number of tissue samples corresponding to four distinct types of small round blue cell tumors.

For each tissue sample, 2308 gene expression measurements are available. 

This data were originally reported in: Khan J, Wei J, Ringner M, Saal L, Ladanyi M, Westermann F, Berthold F, Schwab M, Antonescu C, Peterson C, and Meltzer P. Classification and diagnostic prediction of cancers using gene expression profiling and artificial neural networks. Nature Medicine, v.7, pp.673-679, 2001. 

Here we will consider the subset of the `Khan` dataset where we focus on differences in gene expression between two cancer sub-types: 

+ Rhabdomyosarcoma aka group `2` 

+ Burkitt’s lymphoma aka group `4`

```{r}

library(tidyverse)
library(knitr) # required for nice table output
library(ggthemes) # required for nice colors

options(digits = 3) # number of digits printed by R default (vectors, data.frames, lists)
options(pillar.sigfig = 3) # number of digits printed by tibbles default.

set.seed(42) # to reproduce the example
theme_set(theme_minimal(base_size = 18)) # to get homogeneous figs

df_khan <- read_rds(r"(U18\khan_2vs4.tibble2.rds)")
```

# Looking at gene 11


# The one sample t-test for group `2` against a fixed mean

This example is somewhat artificial unless you have external info to specify a null hypothesis that makes sense: like we know the normal gene expression level of `gene_11` is say 1.3 and we want to test that.

H0: Genes have a expression that is normally distributed with mean=1.3
Ha: Genes have a different expression

```{r}

pd <- df_khan %>% filter(cancer_group == 2)

ggplot(data = pd, mapping = aes(x = gene_11)) +
    geom_histogram() +
    geom_vline(color = "red", xintercept = 1.3) +
    facet_wrap(~cancer_group, ncol = 1) +
    NULL
```

## The mean of expression of gene 11

Remember last week? (no) You found the mean and SE of gene 11.

```{r}

df_khan %>%
    group_by(cancer_group) %>%
    summarize(
        n_obs = n(),
        mean_exp = mean(gene_11),
        std_dev = sd(gene_11),
        std_error = std_dev / sqrt(n_obs)
    ) %>%
    kable(col.names = c("Cancer group", "n", "mean expression", "SD", "SE (mean)")) %>%
    {
        .
    }
```

So we see that the mean expression in group 2 is 1.61.

## Extract values for the test

We extract all observations in group `2` and store it in a new vector

```{r}

group2_values <- df_khan %>%
    filter(cancer_group == "2") %>%
    pull(gene_11)

group2_values
```

Then we use the very versatile function `t.test`

```{r}

myvalues <- c(
    1.29, 0.99, 1.32, 1.59, 1.73, 1.26, 1.3, 1, 1.44, 1.22, 1.44,
    1.49, 1.56, 1.23, 1.45, 1.01, 1.19, 1.18, 0.87, 1.36
)

t.test(x = myvalues)
t.test(x = myvalues, mu = 0)
t.test(x = myvalues, mu = 1.15)
```

## >Q1. Test group2_values against H0: mu=1.3. What is your conclusion?


```{r}

t.test(x = group2_values, mu = 1.3)
```

---

We already knew from the descriptive stats that in group 2, the observed mean was 1.61 with SE=0.0705, so we had a hunch that we would reject H0 (that assumes mu=1.3) and get a quite small p-value.

Note that the 2SEs rule works 

Note we can also save the output of t.test

```{r}

mytest1 <- t.test(x = group2_values, mu = 1.3)
names(mytest1)
mytest1$stderr # check we had already done that
mytest1$method # What kind of test did we do?
mytest1$alternative # check we wanted a two-sided
```

You can extract also the p-value, etc.

# Two sample t-test

```{r}

myvalues1 <- c(
    1.34, 1.26, 1.31, 1.35, 1.31, 1.46, 1.3, 1.31, 1.42, 1.24,
    1.35, 1.19, 1.47, 1.32, 1.3, 1.23, 1.35, 1.27, 1.3, 1.48
)

myvalues2 <- c(
    1.23, 1.14, 1.27, 1.21, 1.31, 1.39, 1.15, 1.3, 1.26, 1.34,
    1.23, 1.33, 1.08, 1.42, 1.34, 1.23, 1.11, 1.31
)

t.test(x = myvalues1, y = myvalues2)
try(t.test(x = myvalues1, y = myvalues2, paired = T)) # Why does it fail? - They have different lengths!
t.test(x = myvalues1, y = myvalues2, var.equal = T)
```

## >Q2. Which of the three test runs is the correct two sample t test, that assumes equal variance in the two groups?

the last one assumes equal variance.

---

## >Q3. Now you are ready - use the t.test() to test group2 vs. group4 for gene 11.

What is your conclusion. Do the two groups have the same expression?

```{r}

group2_values <- df_khan %>%
    filter(cancer_group == "2") %>%
    pull(gene_11)
group4_values <- df_khan %>%
    filter(cancer_group == "4") %>%
    pull(gene_11)

t.test(x = group2_values, y = group4_values, var.equal = T)
```

t.test(x = group2_values, y = group4_values, var.equal = T)                                                                                                                                                                                         
                                                                                                                                                                                                                                                        
Two Sample t-test                                                                                                                                                                                                                               
                                                                                                                                                                                                                                                        
data:  group2_values and group4_values                                                                                                                                                                                                                  
t = -2, df = 52, p-value = 0.04                                                                                                                                                                                                                         
alternative hypothesis: true difference in means is not equal to 0                                                                                                                                                                                      
95 percent confidence interval:
 -0.45777 -0.00971              # what el fuk?
sample estimates:
mean of x mean of y
     1.61      1.85

---

# Non-parametric test

We use the Mann-Whitney test to compare means of two groups.

In R (and many other places) it is also called Wilcoxon Rank Sum test. So we use the function:

wilcox.test()

It uses the same input as the two sample t.test()

```{r}

myvalues1 <- c(
    1.34, 1.26, 1.31, 1.35, 1.31, 1.46, 1.3, 1.31, 1.42, 1.24,
    1.35, 1.19, 1.47, 1.32, 1.3, 1.23, 1.35, 1.27, 1.3, 1.48
)

myvalues2 <- c(
    1.23, 1.14, 1.27, 1.21, 1.31, 1.39, 1.15, 1.3, 1.26, 1.34,
    1.23, 1.33, 1.08, 1.42, 1.34, 1.23, 1.11, 1.31
)

wilcox.test(x = myvalues1, y = myvalues2, exact = F)
```

Warning message:
In wilcox.test.default(x = myvalues1, y = myvalues2, exact = T) :
  cannot compute exact p-value with ties

#### >Q4. Test gene 11 using the wilcox.test().

```{r}
wilcox.test(x = group2_values, y = group4_values, exact = F)
```

data:  group2_values and group4_valuesntinuity correction
W = 265, p-value = 0.09
alternative hypothesis: true location shift is not equal to 0

>What is your conclusion. Do the two groups have the same expression?

they does yes


>If you get a different conclusion - why?

I am indecisive and dyslexic
i mean, i cant re-decide that quick


---

# Permutation test

So - now we are really confused. We don't really know what to trust.

We can quickly check normality

```{r}

group2_values <- df_khan %>%
    filter(cancer_group == "2") %>%
    pull(gene_11)
group4_values <- df_khan %>%
    filter(cancer_group == "4") %>%
    pull(gene_11)

shapiro.test(group2_values)

shapiro.test(group4_values)
```

r$> shapiro.test(group2_values)

        Shapiro-Wilk normality test

data:  group2_values
W = 1, p-value = 0.7


r$>

r$> shapiro.test(group4_values)

        Shapiro-Wilk normality test

data:  group4_values
W = 1, p-value = 0.5

They look ok - so my best guess right now is that the t.test() is ok.

What should we do then? We can use permutation test to get a p value.

For gene 11 we just switch around labels `2` and `4` at "random".

We first make a small dataset and demonstrate the permutation of label 

```{r}

set.seed(0) # Make sure we ALL get the same permutation

(df_khan_2 <- df_khan %>% select(cancer_group, gene_11))

(df_khan_2 <- df_khan_2 %>% mutate(perm_group = sample(x = cancer_group)))
```

In `df_g11`, we have added a new column `perm_group` that contains a permutation of `cancer_group `label (shuffling the group labels)

Let's redo the Welch t-test with these shuffled labels: we just extract the observations matching these shuffled groups: 

To distinguish them from original data we call them `pseudo4` and `pseudo2`

NOTE THE FILTER!! WE use the `perm_group` !!!

```{r}

pseudo2 <- df_khan_2 %>%
    filter(perm_group == "2") %>%
    pull(gene_11)

pseudo4 <- df_khan_2 %>%
    filter(perm_group == "4") %>%
    pull(gene_11)

t.test(x = pseudo2, y = pseudo4)
```

 t.test(x = pseudo2, y = pseudo4)

        Welch Two Sample t-test

data:  pseudo2 and pseudo4
t = -0.3, df = 46, p-value = 0.8
alternative hypothesis: true difference in means is not equal to 0
95 percent confidence interval:
 -0.270  0.205
sample estimates:
mean of x mean of y
     1.71      1.74

Now we can do it a few times with a `for()` loop and output the statistic we get every time

```{r}

group2_values <- df_khan %>%
    filter(cancer_group == "2") %>%
    pull(gene_11)
group4_values <- df_khan %>%
    filter(cancer_group == "4") %>%
    pull(gene_11)

x <- t.test(x = group2_values, y = group4_values, var.equal = T)

observed <- abs(x$statistic)

set.seed(0)

for (i in 1:8) {
    df_khan_2 <- df_khan_2 %>% mutate(perm_group = sample(x = cancer_group))
    pseudo2 <- df_khan_2 %>%
        filter(perm_group == "2") %>%
        pull(gene_11)
    pseudo4 <- df_khan_2 %>%
        filter(perm_group == "4") %>%
        pull(gene_11)

    x <- t.test(x = pseudo2, y = pseudo4)

    cat("t-test t value of permutation ", i, " :", x$statistic, "\n")
}
```


t-test t value of permutation  1  : -0.274 
t-test t value of permutation  2  : 0.434 
t-test t value of permutation  3  : 1.03 
t-test t value of permutation  4  : -0.406
t-test t value of permutation  5  : -0.206 
t-test t value of permutation  6  : -1.99 
t-test t value of permutation  7  : -0.728 
t-test t value of permutation  8  : 0.313

## >Q.5 Adapt the code to do a real permutation based test


```{r}
set.seed(0)

group2_values <- df_khan %>%
    filter(cancer_group == "2") %>%
    pull(gene_11)
group4_values <- df_khan %>%
    filter(cancer_group == "4") %>%
    pull(gene_11)

x <- t.test(x = group2_values, y = group4_values, var.equal = T)

observed <- abs(x$statistic)

t <- rep(NA, 1000)
for (i in 1:length(t)) {
    df_khan_2 <- df_khan_2 %>% mutate(perm_group = sample(x = cancer_group))

    pseudo2 <- df_khan_2 %>%
        filter(perm_group == "2") %>%
        pull(gene_11)

    pseudo4 <- df_khan_2 %>%
        filter(perm_group == "4") %>%
        pull(gene_11)

    t[i] <- abs(t.test(x = pseudo2, y = pseudo4)$statistic)
}
```

```{r}
sum(t >= observed)
```


```{r}
cat("perm-test p-value: ", (2 * (sum(t >= observed) / 1000)), "\n")
```


We want to do 1.000 permutations and, for each shuffle, recompute the t-test statistics,... store the 1.000 statistics in a vector.

Then when you have your permuted t.values you count the number that were more extreme and calculate the p value.

We use the ABSOLUTE t values - since it can be more extreme in two directions (group 2 >> group 4 or group 4 >> group 2)

HINT for calculating p values from permutations:

```{r}

observed <- 3.5
permute_ts <- c(
    -10.08, -10.6, -19.16, -9.42, 7.67, -18.61, -17.49, -4.09,
    -0.4, 5.24, 8.94, -16.93, -3.14, 18.95, 12.55, -11.05, -0.41,
    -19.55, -9.66, -1.82, 9.64, 19.63, -6.71, 17.79, 18.47, 15.97,
    -0.29, 11.4, 12.12, 7.11, 3.25, -6.77, -19.95, -17.37, -16.56,
    -19.48, -6.57, -15.23, 3.74, -18.51, -19.62, -13.54, 13.28, 10.67,
    -9.09, -12.47, -10.97, -17.52, -17.6, -13.38
)

sum(abs(permute_ts) >= abs(observed))

sum(abs(permute_ts) >= abs(observed)) / length(permute_ts)

(sum(abs(permute_ts) >= abs(observed)) + 1) / (1 + length(permute_ts))
```

What is your final conclusion about gene 11? Is it significant or not?

it is significant, they is not the same. 

---

# Increase speed of permutation test

The following is just a few tricks to speed things up. 

Working of tibbles or data frames in for loops is super slow.

There are many many ways you can increase the speed in R.

Now I will show you a few tricks.

## Trick 1 - NO  tibbles inside the for loop:

To see the difference we increase to 10.000 permutations

```{r}

set.seed(0)

g2 <- df_khan_2 %>%
    filter(cancer_group == "2") %>%
    pull(gene_11)
g4 <- df_khan_2 %>%
    filter(cancer_group == "4") %>%
    pull(gene_11)

n1 <- length(g2)
n2 <- length(g4)

permute_ts <- rep(NA, 10000)

for (i in 1:length(permute_ts)) {
    permuted_values <- sample(c(g2, g4)) # Much much faster than before
    # First n1 permuted values are assigned to the pseudo2 group
    pseudo2 <- permuted_values[1:n1]
    # The remaining value are assigned to pseudo 4 group
    pseudo4 <- permuted_values[(1 + n1):(n1 + n2)]
    x <- t.test(x = pseudo2, y = pseudo4, alternative = "two.sided")
    permute_ts[i] <- x$statistic
    if (i %% 100 == 0) {
        cat(".")
    }
}

(observed <- t.test(x = g2, y = g4)$statistic)

(sum(abs(permute_ts) >= abs(observed)) + 1) / (1 + length(permute_ts))
```

## Trick 2 - drop t.test()

Now we also skip the t.test() - since we actually just compare means.

So we build a null distribution of the difference in means.

This is EXACTLY like the original mosquito example.

This is so fast that you can quickly run 1.000.000 permutations

```{r}

set.seed(0)

g2 <- df_khan_2 %>%
    filter(cancer_group == "2") %>%
    pull(gene_11)
g4 <- df_khan_2 %>%
    filter(cancer_group == "4") %>%
    pull(gene_11)

n1 <- length(g2)
n2 <- length(g4)

permute_diff <- rep(NA, 1000000) # NOTE NEW NAME, noted (I lied)

starttime <- Sys.time()

for (i in 1:length(permute_diff)) {
    permuted_values <- sample(c(g2, g4))
    pseudo2 <- permuted_values[1:n1]
    pseudo4 <- permuted_values[(1 + n1):(n1 + n2)]
    permute_diff[i] <- abs(mean(pseudo2) - mean(pseudo4)) # Note abs()
}

endtime <- Sys.time()

endtime - starttime


(observed <- t.test(x = g2, y = g4)$statistic)
```




## >Q.6 What is the estimated p value from 1 million permutations?

NOTE: we are now using the difference in means and our test statistic.
```{r}
cat("perm-test p-value: ", (sum(abs(permute_diff) >= abs(observed)) + 1) / (1 + length(permute_diff)), "\n")
```
---

perm-test p-value:  1e-06 


## >Q7. How long time will it take to permutation test all genes?

```{r}
(NCOL(df_khan) - 1) * (endtime - starttime)
```

a few seconds = 24806 if u is not scrub. cant go lower, prolly can go higher. 

---


# Test all genes

## >Q8. Using all the examples and solutions above: make some code that fill in the table below with correct p values.

```{r}

results <- tibble(
    column = 2:ncol(df_khan),
    gene_name = names(df_khan)[column],
    pval_ttest = NA, # the p value from the t.test
    pval_wilcox = NA
) # the p value from wilcox.test
starttime <- Sys.time()

cg2 <- df_khan %>%
    filter(cancer_group == "2")
cg4 <- df_khan %>%
    filter(cancer_group == "4")

for (i in 1:nrow(results)) {

    # You can get gene_11 with
    gene_name <- pull(results[i, 2])
	
    cg2gene <- cg2 %>% pull(gene_name)
    cg4gene <- cg4 %>% pull(gene_name)

    results[i, 3] <- t.test(x = cg2gene, y = cg4gene)$p.value
    results[i, 4] <- wilcox.test(x = cg2gene, y = cg4gene, exact = F)$p.value

    if (i %% 100 == 0) {
        cat("gene: ", i, " / 2308 \n")
    }
}
endtime <- Sys.time()

endtime - starttime
head(results)
```

```{r}
write_rds(results, r"(U18\Q8.rds)")
```


---


# >Q9. How many genes are significant with BOTH tests?

```{r}
results %>% count(pval_ttest < 0.05 & pval_wilcox < 0.05)
```

  `pval_ttest < 0.05 & .$pval_wilcox < 0.05`     n
  <lgl>                                      <int>
1 FALSE                                       1799
2 TRUE                                         509

---

# >Q10. Are the p values correlated? (plot them)

```{r}

results %>%
    mutate(log_pval_ttest = -log(pval_ttest, base = 10), log_pval_wilcox = -log(pval_wilcox, base = 10)) %>%
    ggplot(aes(x = log_pval_ttest, y = log_pval_wilcox, color = log_pval_ttest)) +
    scale_color_gradientn(colours = rainbow(100)) +
    geom_point() +
    xlim(0, 20) +
    ylim(0, 20) +
    theme(legend.position = "none")
```


---


