---
output:
  html_document:
        theme: readable
editor_options: 
  chunk_output_type: console
---

# Probability and disease

The world is presently facing the worst pandemic in more than 100 years.

All numbers presented here are completely fictional, but illustrate some points about disease transmission and the effect of vaccines.


```{r}

library(tidyverse)

```

Let's assume that in a study we find that 1200 occurences of close contact between a infected and uninfected individual resulted in 312 new infections.

>Q1.1 What is the estimated probability of being infected if you are in close contact with an infectious individual? 

```{r}
312 / 1200
```
pr=0.26
Example of answer formating :

```{r}
(probability <- 0.00000001234)
(probability <- 1.234e-08)
(probability <- 40 / 97)
```


## Vaccines

Vaccines are known to give partial immunity against many diseases, but rarely perfect protection.

Now we assume the following:

1. The  probablity of being infected upon close contact is 0.25

2. If you are vaccinated against the disease, the probability drops to 0.10

```{r}

p1 <- 0.25 # probability of infection pr. contact when not vaccinated
p2 <- 0.10 # probability of infection pr. contact when vaccinated
```

>Q1.2 What is probability of NOT becoming infected if you are in close contact with a single infectious individual and you are NOT vaccinated?

```{r}
1 - 0.25
```
1-0.25=0.75
>Q1.3 What is probability of NOT becoming infected if you are in close contact with a single infectious individual and you ARE vaccinated?
```{r}
1 - 0.10
```
0.9
>Q1.4 What is probability of becoming infected if you are in close contact with 5 infectious individuals and you are NOT vaccinated? 

```{r}
1 - (0.75^5)
```
0.7626953
>Q1.5 What is probability of becoming infected if you are in close contact with 5 infectious individuals and you ARE vaccinated? 


```{r}
1 - (0.9^5)
```
0.40951

## Now we assume that YOU are infected.

Now we turn things around a little bit and look at your role as an infectious individual.

Let us assume that you meet 20 people on a day where you are actually infectious (but asymptomatic).

The probilities of infecting another person when in close contact is still p1 and p2.

>Q1.6 What is probability that you infect 0 of the 20 people (none of them are vaccinated)?

```{r}
0.75^20
```
0.003171212
>Q1.7 What is probability that you infect 5 or more of the 20 people (none of them are vaccinated)?

```{r}
1 - pbinom(q = 4, size = 20, prob = 0.25)
pbinom(q = 4, size = 20, prob = 0.25, lower.tail = F)
```
0.5851585

>Q1.8 What is probability that you infect 5 or more of the 20 people (ALL of them are vaccinated)?

```{r}
1 - pbinom(q = 4, size = 20, prob = 0.10)
```
0.0431745

## Are you a superspreader

Some infected people are known to infect more easily than others. 

Now, lets imagine you attended two parties on two different days and later you test positive for the disease.

We also assume the following:

* No other people at the parties were infected. So all new infections were caused by you.

* You had close contact with everyone at the party. So you could potentially have infected everyone.

Now, please test the null hypothesis that you are a "normal" infectious person. The alternative hypothesis is that your probability of infecting another person is different from the hypothetical one.

>Q1.9 What is the p value if you infected 27 guests at a party with 76 unvaccinated guests in total.

```{r}
binom.test(x = 27, n = 75, p = 0.25)

```
p-value = 0.03265
 Exact binomial test

data:  27 and 75
number of successes = 27, number of trials = 75, p-value = 0.03265
alternative hypothesis: true probability of success is not equal to 0.25
95 percent confidence interval:
 0.2523224 0.4791016
sample estimates:
probability of success
                  0.36
Hint: The 76 guest are you and 75 others.


>Q1.10 What is the p value if you infected 7 guests at a party with 55 vaccinated guests in total.

```{r}
binom.test(x = 7, n = 54, p = 0.1)
```
p-value = 0.4916
Hint: The 55 guest are you and 54 others. All the othes were vaccinated






