---
output:
  html_document:
        theme: readable
editor_options: 
  chunk_output_type: console
---

# Probability and disease

The world is presently facing the worst pandemic in more than 100 years.

All numbers presented here are completely fictional, but illustrate some points about disease transmission and the effect of vaccines.


```{r}

library(tidyverse)

```

Let's assume that in a study we find that 1200 occurences of close contact between a infected and uninfected individual resulted in 312 new infections.

>Q1.1 What is the estimated probability of being infected if you are in close contact with an infectious individual? 

Example of answer formating :

```{r}
(probability <- 0.00000001234)
(probability <- 1.234e-08)
(probability <- 40/97)

```

---------!begin suggested answer ------------------------

We calculate the probability

```{r}

answer <- tibble()

(x <- 322/1200)

answer <- rbind(answer,tibble(
  question    = "1", 
  true_answer = x, 
  type        = "numeric",
  points      = 1,
  tolerance   = 0.01))

```

---------!end suggested answer ------------------------

## Vaccines

Vaccines are known to give partial immunity against many diseases, but rarely perfect protection.

Now we assume the following:

1. The  probablity of being infected upon close contact is 0.25

2. If you are vaccinated against the disease, the probability drops to 0.10

```{r}

p1 <- 0.25 # probability of infection pr. contact when not vaccinated
p2 <- 0.10 # probability of infection pr. contact when vaccinated

```

>Q1.2 What is probability of NOT becoming infected if you are in close contact with a single infectious individual and you are NOT vaccinated?

---------!begin suggested answer ------------------------

```{r}

(x <- 1-p1)

answer <- rbind(answer,tibble(
  question    = "2", 
  true_answer = x, 
  type        = "numeric",
  points      = 1,
  tolerance   = 0.001))

```
---------!end suggested answer ------------------------

>Q1.3 What is probability of NOT becoming infected if you are in close contact with a single infectious individual and you ARE vaccinated?

---------!begin suggested answer ------------------------

```{r}

(x <- 1-p2)

answer <- rbind(answer,tibble(
  question    = "3", 
  true_answer = x, 
  type        = "numeric",
  points      = 1,
  tolerance   = 0.001))

```

---------!end suggested answer ------------------------

>Q1.4 What is probability of becoming infected if you are in close contact with 5 infectious individuals and you are NOT vaccinated? 

---------!begin suggested answer ------------------------

It is easiest to calculate to probability of NOT becoming infected at all.

Then to say 1-that.

```{r}

p_not_infected <- (1-p1)^5
x <- 1 - p_not_infected

answer <- rbind(answer,tibble(
  question    = "4", 
  true_answer = x, 
  type        = "numeric",
  points      = 3,
  tolerance   = 0.01))

```

---------!end suggested answer ------------------------

>Q1.5 What is probability of becoming infected if you are in close contact with 5 infectious individuals and you ARE vaccinated? 

---------!begin suggested answer ------------------------

It is easiest to calculate to probability of NOT becoming infected at all.

Then to say 1-that.

Because you may actually be infected by more than one of the infectious individuals.

```{r}

p_not_infected <- (1-p2)^5
x <- 1 - p_not_infected

answer <- rbind(answer,tibble(
  question    = "5", 
  true_answer = x, 
  type        = "numeric",
  points      = 3,
  tolerance   = x/50))

```

---------!end suggested answer ------------------------

## Now we assume that YOU are infected.

Now we turn things around a little bit and look at your role as an infectious individual.

Let us assume that you meet 20 people on a day where you are actually infectious (but asymptomatic).

The probilities of infecting another person when in close contact is still p1 and p2.

>Q1.6 What is probability that you infect 0 of the 20 people (none of them are vaccinated)?

---------!begin suggested answer ------------------------

It is easiest to calculate to probability of NOT infection anyone, since there is only one combination where that happens.

```{r}

(x <- (1-p1)^20)

answer <- rbind(answer,tibble(
  question    = "6", 
  true_answer = x, 
  type        = "numeric",
  points      = 3,
  tolerance   = x/50))

```

---------!end suggested answer ------------------------

>Q1.7 What is probability that you infect 5 or more of the 20 people (none of them are vaccinated)?

---------!begin suggested answer ------------------------

You can infect 5,6,7..20 people in total.

So we need to use the binomial distribution.

```{r}

p_4_or_less <- pbinom(q = 4, size = 20, prob = p1, lower.tail = T)

pbinom(q = 4, size = 20, prob = p1, lower.tail = T)
# This gives p for 0,1,2,3,4 
# From ?pbinom
# lower.tail	logical; if TRUE (default), probabilities are P[X ≤ x], otherwise, P[X > x].

x <- 1-p_4_or_less

# Another way is simply to say
(pbinom(q = 4, size = 20, prob = p1, lower.tail = F))

answer <- rbind(answer,tibble(
  question    = "7", 
  true_answer = x, 
  type        = "numeric",
  points      = 3,
  tolerance   = x/50))

```

---------!end suggested answer ------------------------

>Q1.8 What is probability that you infect 5 or more of the 20 people (ALL of them are vaccinated)?

---------!begin suggested answer ------------------------

You can infect 5,6,7..20 people in total.

So we need to use the binomial distribution.

```{r}

(x <- pbinom(q = 4, size = 20, prob = p2, lower.tail = F))

answer <- rbind(answer,tibble(
  question    = "8", 
  true_answer = x, 
  type        = "numeric",
  points      = 3,
  tolerance   = x/50))

```

---------!end suggested answer ------------------------

## Are you a superspreader

Some infected people are known to infect more easily than others. 

Now, lets imagine you attended two parties on two different days and later you test positive for the disease.

We also assume the following:

* No other people at the parties were infected. So all new infections were caused by you.

* You had close contact with everyone at the party. So you could potentially have infected everyone.

Now, please test the null hypothesis that you are a "normal" infectious person. The alternative hypothesis is that your probability of infecting another person is different from the hypothetical one.

>Q1.9 What is the p value if you infected 27 guests at a party with 76 unvaccinated guests in total.

Hint: The 76 guest are you and 75 others.

---------!begin suggested answer ------------------------

We need to test if your probability really is 0.25

The maximum number you could infect is 75

```{r}

(result <- binom.test(x = 27, n = 75, p = 0.25))

x <- result$p.value

answer <- rbind(answer,tibble(
  question    = "9", 
  true_answer = x, 
  type        = "numeric",
  points      = 3,
  tolerance   = x/50))

```

You are a super spreader.

---------!end suggested answer ------------------------

>Q1.10 What is the p value if you infected 7 guests at a party with 55 vaccinated guests in total.

Hint: The 55 guest are you and 54 others. All the othes were vaccinated

---------!begin suggested answer ------------------------

We need to test if your probability really is 0.15

The maximum number you could infect is 54

```{r}

(result <- binom.test(x = 7, n = 54, p = 0.15))

x <- result$p.value

answer <- rbind(answer,tibble(
  question    = "10", 
  true_answer = x, 
  type        = "numeric",
  points      = 3,
  tolerance   = x/50))

```

You are not a superspreader when people are vaccinated.

---------!end suggested answer ------------------------



---------!begin suggested answer ------------------------

## Hard coded information

```{r}

topic             <- "t1"  # If you change this to e.g. 2 - you should search & replace all >Q1. with >Q2.
topic_description <- "# Disease transmission"
file_description  <- "disease_transmission"

```

## Answers table

```{r}

answer

```


## Saving exam files

```{r}

answer$question   <- paste(topic, ".", answer$question, sep="") # Reformat column

# Save answer sheet for students to fill out (including points)
x <- paste("answer$", answer$question, " = NA # (",answer$points," points)",sep="")

answer_filename <- paste("exam_answer_sheet.",topic, ".", file_description, ".txt", sep="")
write_lines(topic_description, path = answer_filename)
write_lines(x, path =answer_filename, append = T)

# Save answer key for grading
key_filename <- paste("answer.key.",topic, ".", file_description, ".tsv", sep="")
write_tsv(x = answer, path = key_filename)

```

---------!end suggested answer ------------------------


