---
output:
  html_document:
        theme: readable
editor_options: 
  chunk_output_type: console
---

# Deaths in Denmark, 1981 and 2018

We have downloaded data in early august from statistikbanken in the table DOD1

Url: http://www.statistikbanken.dk/DOD1

We are looking at sex differences and changes from 1981 to 2018. 

The data consist of 3 columns:

 * the sex of the deceased person
 * the cause of death
 * the year of death
 
We have chosen to focus on 3 different causes of death only:

 * Mental disorders
 * Motor vehicle accidents
 * Suicide and self inflicted injury
 
```{r message=FALSE, warning=FALSE}

library(tidyverse)

df <- read_tsv(file = "t2.deaths_in_Denmark_dataset.tsv")

```

>Q2.1 How many deaths from these 3 causes were there in total in 1981 in this dataset?


>Q2.2 How many men died from these 3 causes in 2018? 


>Q2.3 What proportion of deaths from Motor vehicle accidents in 1981 were men?

Hint: Proportion is a number in the range [0;1]


>Q2.4 How many male deaths from Motor vehicle accidents would you expect in 2018 if both males and females had the same risk?

Hint: We assume that the population has a sex ratio of 50:50, i.e. same number of males and females.


## How to convert our data to a table

```{r}

tmp <- df %>% 
  filter(cause=="Mental disorders") %>% 
  select(sex,year) %>%
  table()

tmp

```

>Q2.5 Test if sex and year are independent for Motor vehicle accidents. Report the p value.


>Q2.6 What is your conclusion about motor vehicle accidents?

Answer should be a number indicating one of the following statements:

1. Sex and year are not significantly associated.  
2. Sex and year are weakly significantly associated.  
3. Sex and year are strong significantly associated.  
4. It is not possible to conclude anything.


>Q2.7 Test if cause of death and sex are independent in 1981. Report the p value.


>Q2.8 What is your conclusion?

Answer should be a number indicating one of the following statements:

1. The cause of deaths is not associated to sex.  
2. The cause of deaths is weakly associated to sex.  
3. The cause of deaths is higly associated to sex.  
4. It is not possible to conclude anything.


>Q2.9 Test if cause of death and sex are independent in 2018 Report the p value.


>Q2.10 What is the sample size in 2018?


>Q2.11 What are the degrees of freedom for the last test you did?


## Comparing associations

We can’t directly say if the association between cause of death and sex is strongest in 1981 or 2018, since we have different sample sizes for the two years.

Because of differences in samples size, we expect the year with highest sample size to have higher chi square values.

If we want to find the "strongest" effect, we need to correct for this difference in sample size.

A way of correcting for sample size is by calculating the phi coefficient. You have never heard of it before, so here comes the simple definition:

Phi^2 is defined as the chi square value divided by the sample size. 

Phi is then the square root of the Phi^2 value.

Definition of phi: https://en.wikipedia.org/wiki/Phi_coefficient

In statistics, the phi coefficient (also called mean square contingency coefficient) is a measure of association for two binary variables.

>Q2.12 What is the phi coefficient for 1981?


>Q2.13 What is the phi coefficient for 1981?


>Q2.14 What is your final conclusion?

Answer should be a number indicating one of the following statements:

1. It is not possible to conclude anything.
2. The association has decreased from 1981 to 2018.
3. The association has increased from 1981 to 2018.
4. The association is the same in 1981 and 2018



