---
output:
  html_document:
        theme: readable
editor_options: 
  chunk_output_type: console
---

# Deaths in Denmark, 1981 and 2018

We have downloaded data in early august from statistikbanken in the table DOD1

Url: http://www.statistikbanken.dk/DOD1

We are looking at sex differences and changes from 1981 to 2018. 

The data consist of 3 columns:

 * the sex of the deceased person
 * the cause of death
 * the year of death
 
We have chosen to focus on 3 different causes of death only:

 * Mental disorders
 * Motor vehicle accidents
 * Suicide and self inflicted injury
 
```{r message=FALSE, warning=FALSE}

library(tidyverse)

df <- read_tsv(file = "t2.deaths_in_Denmark_dataset.tsv")

```

>Q2.1 How many deaths from these 3 causes were there in total in 1981 in this dataset?

---------!begin suggested answer ------------------------

```{r}

answer <- tibble()

df %>% count(year)
x <- sum(df$year==1981)

answer <- rbind(answer,tibble(
  question    = "1", 
  true_answer = x, 
  type        = "numeric",
  points      = 1,
  tolerance   = 0.1))

```
---------!end suggested answer ------------------------

>Q2.2 How many men died from these 3 causes in 2018? 

---------!begin suggested answer ------------------------
```{r}

df %>% count(sex,year)

x <- sum(df$sex=="Men" & df$year==2018)

answer <- rbind(answer,tibble(
  question    = "2", 
  true_answer = x, 
  type        = "numeric",
  points      = 1,
  tolerance   = 0.1))

```
---------!end suggested answer ------------------------

>Q2.3 What proportion of deaths from Motor vehicle accidents in 1981 were men?

Hint: Proportion is a number in the range [0;1]

---------!begin suggested answer ------------------------
```{r}

df %>% filter(year==1981) %>% count(sex,year, cause)

men <- sum(df$sex=="Men" & df$year==1981 & df$cause=="Motor vehicle accidents")
women <- sum(df$sex=="Women" & df$year==1981 & df$cause=="Motor vehicle accidents")

x <- men / (men+women)

answer <- rbind(answer,tibble(
  question    = "3", 
  true_answer = x, 
  type        = "numeric",
  points      = 2,
  tolerance   = 0.01))

```
---------!end suggested answer ------------------------

>Q2.4 How many male deaths from Motor vehicle accidents would you expect in 2018 if both males and females had the same risk?

Hint: We assume that the population has a sex ratio of 50:50, i.e. same number of males and females.

---------!begin suggested answer ------------------------

```{r}

men <- sum(df$sex=="Men" & df$year==2018 & df$cause=="Motor vehicle accidents")
women <- sum(df$sex=="Women" & df$year==2018 & df$cause=="Motor vehicle accidents")

x <- mean(c(men,women))

answer <- rbind(answer, tibble(
  question    = "4", 
  true_answer = x, 
  type        = "numeric",
  points      = 3,
  tolerance   = x/100)
  )

```

---------!end suggested answer ------------------------

## How to convert our data to a table

```{r}

tmp <- df %>% 
  filter(cause=="Mental disorders") %>% 
  select(sex,year) %>%
  table()

tmp

```

>Q2.5 Test if sex and year are independent for Motor vehicle accidents. Report the p value.

---------!begin suggested answer ------------------------

This is a simple chi square test of independence.

```{r}
unique(df$cause)

tmp <- df %>% 
  filter(cause=="Motor vehicle accidents") %>% 
  select(sex,year) %>%
  table()

result <- chisq.test(tmp, correct = F)

x <- result$p.value

answer <- rbind(answer, tibble(
  question    = "5", 
  true_answer = x, 
  type        = "numeric",
  points      = 3,
  tolerance   = x/10)
  )

```

---------!end suggested answer ------------------------

>Q2.6 What is your conclusion about motor vehicle accidents?

Answer should be a number indicating one of the following statements:

1. Sex and year are not significantly associated.  
2. Sex and year are weakly significantly associated.  
3. Sex and year are strong significantly associated.  
4. It is not possible to conclude anything.

---------!begin suggested answer ------------------------

```{r}

answer <- rbind(answer, tibble(
  question    = "6", 
  true_answer = 1, 
  type        = "numeric",
  points      = 2,
  tolerance   = 0.1)
  )

```

---------!end suggested answer ------------------------

>Q2.7 Test if cause of death and sex are independent in 1981. Report the p value.

---------!begin suggested answer ------------------------

This is a simple chi square test of independence.

```{r}

tmp <- df %>% 
  filter(year==1981) %>% 
  select(cause,sex) %>%
  table()

result <- chisq.test(tmp, correct = F)

x <- result$p.value

answer <- rbind(answer, tibble(
  question    = "7", 
  true_answer = x, 
  type        = "numeric",
  points      = 3,
  tolerance   = x/10)
  )

```

---------!end suggested answer ------------------------

>Q2.8 What is your conclusion?

Answer should be a number indicating one of the following statements:

1. The cause of deaths is not associated to sex.  
2. The cause of deaths is weakly associated to sex.  
3. The cause of deaths is higly associated to sex.  
4. It is not possible to conclude anything.

---------!begin suggested answer ------------------------

```{r}

answer <- rbind(answer, tibble(
  question    = "8", 
  true_answer = 1, 
  type        = "numeric",
  points      = 2,
  tolerance   = 0.1)
  )

```

---------!end suggested answer ------------------------

>Q2.9 Test if cause of death and sex are independent in 2018 Report the p value.

---------!begin suggested answer ------------------------

This is a simple chi square test of independence.

```{r}

tmp <- df %>% 
  filter(year==2018) %>% 
  select(cause,sex) %>%
  table()

result <- chisq.test(tmp, correct = F)

x <- result$p.value

answer <- rbind(answer, tibble(
  question    = "9", 
  true_answer = x, 
  type        = "numeric",
  points      = 3,
  tolerance   = 1e-10)
  )

```

---------!end suggested answer ------------------------

>Q2.10 What is the sample size in 2018?

---------!begin suggested answer ------------------------

```{r}

x <- sum(df$year==2018)

answer <- rbind(answer, tibble(
  question    = "10", 
  true_answer = x, 
  type        = "numeric",
  points      = 2,
  tolerance   = 0.1)
  )

```

---------!end suggested answer ------------------------

>Q2.11 What are the degrees of freedom for the last test you did?

---------!begin suggested answer ------------------------

```{r}

answer <- rbind(answer, tibble(
  question    = "11", 
  true_answer = 2, 
  type        = "numeric",
  points      = 2,
  tolerance   = 0.1)
  )

```

---------!end suggested answer ------------------------

## Comparing associations

We can’t directly say if the association between cause of death and sex is strongest in 1981 or 2018, since we have different sample sizes for the two years.

Because of differences in samples size, we expect the year with highest sample size to have higher chi square values.

If we want to find the "strongest" effect, we need to correct for this difference in sample size.

A way of correcting for sample size is by calculating the phi coefficient. You have never heard of it before, so here comes the simple definition:

Phi^2 is defined as the chi square value divided by the sample size. 

Phi is then the square root of the Phi^2 value.

Definition of phi: https://en.wikipedia.org/wiki/Phi_coefficient

In statistics, the phi coefficient (also called mean square contingency coefficient) is a measure of association for two binary variables.

>Q2.12 What is the phi coefficient for 1981?

---------!begin suggested answer ------------------------

We get the chi square from the test

```{r}

tmp <- df %>% 
  filter(year==1981) %>% 
  select(cause,sex) %>%
  table()

result <- chisq.test(tmp, correct = F)

result$statistic
sum(tmp)

phi1981 <- sqrt(result$statistic / sum(tmp))
x <- phi1981

answer <- rbind(answer, tibble(
  question    = "12", 
  true_answer = x, 
  type        = "numeric",
  points      = 3,
  tolerance   = x/10)
  )

```

---------!end suggested answer ------------------------

>Q2.13 What is the phi coefficient for 1981?

---------!begin suggested answer ------------------------

We get the chi square from the test

```{r}

tmp <- df %>% 
  filter(year==2018) %>% 
  select(cause,sex) %>%
  table()

result <- chisq.test(tmp, correct = F)

result$statistic
sum(tmp)

phi2018 <- sqrt(result$statistic / sum(tmp))
x <- phi2018

answer <- rbind(answer, tibble(
  question    = "13", 
  true_answer = x, 
  type        = "numeric",
  points      = 3,
  tolerance   = x/10)
  )

```

---------!end suggested answer ------------------------

>Q2.14 What is your final conclusion?

Answer should be a number indicating one of the following statements:

1. It is not possible to conclude anything.
2. The association has decreased from 1981 to 2018.
3. The association has increased from 1981 to 2018.
4. The association is the same in 1981 and 2018

---------!begin suggested answer ------------------------

phi2018 is bigger than for 1981.

It is approximately twice as big.

That would indicated the association is a lot stronger in 2018 than 1981 - so it has INCREASED

```{r}

phi2018
phi1981

phi2018/phi1981 

answer <- rbind(answer, tibble(
  question    = "14", 
  true_answer = 3, 
  type        = "numeric",
  points      = 2,
  tolerance   = 0.1)
  )

```

---------!end suggested answer ------------------------

---------!begin suggested answer ------------------------

## Hard coded information

```{r}

topic             <- "t2"
topic_description <- "# Deaths in Denmark"
file_description  <- "deaths_in_denmark"

answer

```

## Saving exam files

```{r}

answer$question   <- paste(topic, ".", answer$question, sep="") # Reformat column

# Save answer sheet for students to fill out (including points)
x <- paste("answer$", answer$question, " = NA # (",answer$points," points)",sep="")

answer_filename <-  paste("exam_answer_sheet.",topic, ".", file_description, ".txt", sep="")
write_lines(topic_description, path = answer_filename)
write_lines(x, path =answer_filename, append = T)

# Save answer key for grading
key_filename <- paste("answer.key.",topic, ".", file_description, ".tsv", sep="")
write_tsv(x = answer, path = key_filename)

```

---------!end suggested answer ------------------------

