---
title: "Project A (2)"
output: 
  html_document:
    code_folding: show
    theme: cosmo
    toc: yes
editor_options: 
  chunk_output_type: console
---


# Book exercises

>Do the practice problem 11 in chapter 9: "alcohol and heart disease".


```{r}

```

# Learning goals for this exercise

We start by doing some work on single SNPs to get a good understanding of the calculations

What you will do:

 * Calculate chisquare statitics by hand and use chisquare distribution to get a p value
 * Use R's chi square test function to get a p value
 * use R's fishers exact test to get a p value
 * Compare all the results
 * Compare the time used for test

## Working on 280.000 SNPs

 * Perform a GWAS (280.000 SNPs)
 * Identify which chromosome has the highest fraction of significant SNPs (at alfa=0.005)

# R functions used this week

 * basic math like +,-,/,*
 * for() loops
 * chisq.test()
 * fisher.test()
 * nrow()

# Initialise session
 
```{r}

library(tidyverse)

options(digits = 7)        # number of digits printed by R default (vectors, data.frames, lists)
options(pillar.sigfig = 7) # number of digits printed by tibbles default.

```

# Reading data

```{r}

snpdata <- read_rds(file = "gwas_allele_counts.julia.2008.rds")

```

# Working on a single SNP - number 2059

## Chi.square calculations

We will look at SNP number 2059

First some example code

```{r}

Observed <- snpdata$contingency_table[[2059]]

a <- Observed[1,1]
b <- Observed[1,2]
c <- Observed[2,1]
d <- Observed[2,2]

n <- a+b+c+d

Pr_row1 <- (a+b)/n 
Pr_row2 <- (c+d)/n
Pr_col1 <- (a+c)/n
Pr_col2 <- (b+d)/n

E_a <- Pr_row1 * Pr_col1 * n 
E_b <- Pr_row1 * Pr_col2 * n 
E_c <- Pr_row2 * Pr_col1 * n 
E_d <- Pr_row2 * Pr_col2 * n 

# This is one of several way you can make your own 2x2 table in R.
Expected <- as.table(rbind(c(E_a, E_b), 
                           c(E_c, E_d)))

rownames(Expected) <- rownames(Observed)
colnames(Expected) <- colnames(Observed)

```

```{r}

Observed

Expected

addmargins(Observed)
addmargins(Expected)

```

## calculating stuff

First: you can easily subtract and manipulate two tables

```{r}

Observed-Expected

(Observed-Expected)^2

sum(Observed-Expected)

```

>Q: calculate the chi square value.


---

## P value from a chisquare value and the chi square distribution

Compare the value with the chi square NULL distribution

We use the formula in chapter 9 to calculate the degrees of freedom.

The chi square NULL distribution is different for different degrees of freedom.

For a 2x2 table it's 1*1=1

Here I say that we observed a chisquare value of 5.0

```{r}

chisquare <- 5.0

df <- (nrow(Observed)-1)*(ncol(Observed) - 1) 

pchisq(q = chisquare, df = df, lower.tail = FALSE) 

```

>Q: Get the p value of your SNP?


---

## Using the builtin test in R for comparison

```{r}

x <- chisq.test(Observed, correct = F)
x

# You can get observed and expected counts back
addmargins(x$observed)
addmargins(x$expected)

```

>Q: Compare the values of chi square and the p value and the degrees of freedom for the two approaches

## Yates correction for continuity

>Q: Try and perform a chisquare test with correction for contunuity

Yes - we want you to calculate the modified chi square value and use pchisq() etc.

Also do it with the builtin chisq.test() and compare the results


---

## Simulating the NULL distribution

The chisq.test() can get you simulation results directly. 

Here I run it with 10 and 100000 simulations.

Why should you do it? It works - even if the assumptions are violated.

Why should you NOT do it? It takes longer time.

How does it calculate the p.value? 

It takes the number of simulated chi square values that are >= the observed chi square value anmd call it n_extreme.

Then it assumes the next simulation would be more extreme as well.

p.val = (n.extreme + 1) / (number of simulations + 1)

The problems is: if you want high precision and small p values - then you need millions of simulations.


```{r}

chisq.test(Observed, correct = F, simulate.p.value = TRUE, B = 9)

chisq.test(Observed, correct = F, simulate.p.value = TRUE, B = 300000)

x <- chisq.test(Observed, correct = F, simulate.p.value = TRUE, B = 300000)
x
x$p.value
x$statistic

```

## Compare with builtin function with and without correction

```{r}

chisq.test(Observed, correct = F)

chisq.test(Observed, correct = T)

```

## Compare with fishers exact test

```{r}

fisher.test(Observed)

x <- fisher.test(Observed)

x$p.value

```

Assume that the fisher exact test is the most correct.

>Q: Which of the chisq.test() methods above gave the most correct p value?

```{r}

x <- fisher.test(Observed)
x$p.value

x <- chisq.test(Observed, correct = T)
x$p.value

x <- chisq.test(Observed, correct = F)
x$p.value

x <- chisq.test(Observed, correct = F, simulate.p.value = TRUE, B = 300000)
x$p.value

```

## Soon you will do 280000 test... so you have to worry about speed.

So, the fisher.test() is quite slow.

We will NOT do any timing of the running time. Just believe me.

So, based on your results:

>Q: What settings/strategy for for the chisq.test() will you use for all 280.000 SNPS?

# Working on 280.000 SNPs

## Restart R 

This will clear your memory.

Go to Session>Restart R

Or press ctrl + shift + f10 (windows/linux) or Command+Shift+F10 (mac)

You can also clear the console (ctrl+L)

To see all shortcuts Help>Keyboard shortcuts help

## Reading data (again)

```{r}

library(tidyverse)
snpdata <- read_rds(file = "gwas_allele_counts.julia.2008.rds")

```

## Perform a gwas

Last week you calculated Odds ratios for many snps.

>Q: Use the code above and from last week to test all 280.000 SNPs using the chisq.test
>for each test you want to extract the p value
>and finally save it in column snpdata$p.value

So you should add the p.value as a column to the dataset.

It should run pretty fast, since we use the fast chisq.test()

More precise, but slower: fisher.test()

BONUS task if you have a lot of time: run different kinds of test and compare results


---


## The Manhattan plot

A Manhattan plot is really just a scatterplot with genome position on x and -log10(p) on y.

>Q: Make a manhattan plot of your chisq.test() p values.

Bonus: you can also add chromosome as color to see the different chromosomes.


---

>Q: How many SNPs are significant at alpha = 0.05?



---


>Q: How many did you expect at alpha=0.05?



---

>Q: How many SNPs are significant after bonferroni correction? (interleaf 8)


---

>Q: Which chromosome has the highest percentage of significant SNPs at the 0.005 level?

Please note the 0.005 level!!!

How is this compared to your findings last week?


---

